---
title: "Weekly Assignment #7"
author: "Your Name Here"  # Replace with your name
date: 03/19/2026
date-format: long
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

::: callout-important
## Submission Instructions

1.  Download this `hw7.qmd` file from bCourse.
2.  Save the file in your assignments folder.
3.  Follow the instructions below to complete the assignment. Show your work for all calculations where necessary!
4.  When you are done, click the "Render" button at the top of Positron to generate an .html file.
5.  Submit both `hw7.qmd` and `hw7.html` to bCourse before the deadline.
6.  Make sure your name appears in the "author" field in the YAML header above.
:::

# Instructions

Complete all questions below, writing your code in the code chunks provided. Replace `# Your code here` with your actual code. For questions requiring hand calculations, show your work in the text sections. Make sure your document renders without errors before submitting.

## Setup

```{r}
#| label: setup
#| warning: false
#| message: false

# Clear environment
# ---------
rm(list = ls())

# load libraries
# ---------
library(tidyverse)
library(here)
```

------------------------------------------------------------------------

## Question 1: Estimation Concepts

**a) What does it mean for an estimator to be *unbiased*? What does *efficiency* refer to? In your answer, connect each property to the terms accuracy and precision.**

*Your answer here:*

**b) A researcher reports: "The 95% confidence interval for the proportion of Americans who support stricter gun laws is \[0.51, 0.59\]." A classmate says: "There is a 95% probability that the true proportion falls in that range." Is this correct? Explain in 2–3 sentences.**

*Your answer here:*

**c) You are estimating a population mean from a sample of** $n = 12$. Should you use $z^*$ or $t^*$ as your critical value? Why? What additional assumption is required?

*Your answer here:*

------------------------------------------------------------------------

## Question 2: Confidence Intervals for a Mean

A sociologist collects a random sample of **18 recent immigrants** to the United States and asks how many years they have lived in the country. The sample mean is **8.4 years** and the sample standard deviation is **4.2 years**.

**a) Report the values of** $n$, $\bar{x}$, $s$, and $df$ for this sample.

*Your answer here:*

**b) Use R to find the t\* critical value for a 95% confidence interval with the appropriate degrees of freedom. How does this compare to the z\* critical value of 1.96?**

```{r}

# Your code here

```

*Your answer here:*

**c) Calculate the standard error** $SE(\bar{x})$. Show your work.

```{r}

# Your code here

```

*Your answer here:*

**d) Calculate and interpret the 95% confidence interval.**

```{r}

# Your code here

```

*Your answer here:*

**e) Would the confidence interval be wider or narrower if you had sampled 50 immigrants instead of 18? Explain briefly.**

*Your answer here:*

------------------------------------------------------------------------

## Question 3: Comparing Confidence Intervals Across Groups

Using a larger follow-up sample of **35 immigrants**, the researcher splits respondents by age of arrival: those who **arrived before age 18** (childhood arrivals, $n_1 = 20$, $\bar{x}_1 = 11.4$ yrs, $s_1 = 3.6$ yrs) and those who **arrived at age 18 or older** (adult arrivals, $n_2 = 15$, $\bar{x}_2 = 5.2$ yrs, $s_2 = 4.0$ yrs).

**a) Calculate a 95% confidence interval for the mean years in the U.S. for each group.**

```{r}

# Your code here

```

*Your answer here:*

**b) Do the two confidence intervals overlap? Based on this comparison alone, what can you speculate about differences in mean years in the U.S. between the two groups?**

*Your answer here:*

------------------------------------------------------------------------

::: callout-tip
## Before You Submit

-   [ ] Your name is in the author field at the top of this document
-   [ ] All code chunks run without errors
-   [ ] You've answered all written response questions and shown your work
-   [ ] You've clicked "Render" and successfully generated an HTML file
-   [ ] You're submitting both the .qmd and .html file to bCourse
:::