---
title: "Week 16"
subtitle: "Sociology 106: Quantitative Sociological Methods"
date: 05/05/2026
date-format: long
format:
  revealjs:
    slide-number: true
    chalkboard:
      buttons: false
    preview-links: auto
    css: slides.css
    scrollable: true
    footer: '[GitHub course repo](https://www.kaseyzapatka.com/soc106)'
---

```{r}
#| echo: false
#| results: false
#| warning: false
#| message: false

library(tidyverse)

options(scipen = 999)

interaction_example <- crossing(
  education = seq(8, 16, by = 2),
  group = c("Group A", "Group B")
) |>
  mutate(
    expected_y = case_when(
      group == "Group A" ~ 12 + 1.2 * education,
      group == "Group B" ~ 20 + 0.4 * education
    )
  )

interaction_plot <- ggplot(
  interaction_example,
  aes(x = education, y = expected_y, color = group)
) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2.6) +
  scale_color_manual(values = c("Group A" = "#4E79A7", "Group B" = "#E15759")) +
  labs(
    x = "Education",
    y = "Expected outcome",
    color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

## Agenda {.smaller}

- Final exam format and logistics
- The major concept areas you should be ready for
- How to study efficiently during the final week without trying to memorize everything

::: {.callout-note icon=false}
## Big picture
This review is intentionally high-level. The goal is to help you organize your studying and recognize the kinds of reasoning the exam will ask for, not to preview exact questions.
:::


## Final Exam Format {.smaller}

- **Date:** Monday, May 11, 2026
- **Time:** 11:30 AM to 2:30 PM
- **Place:** Dwinelle 246 (Here)
- **Format:** closed-book, closed-computer
- **Structure:** 10 multiple-choice questions and 10 short-answer questions
- **Points:** 20 points total
- **What you'll get:** all output you need is provided for you

::: {.callout-tip icon=false}
## What this means for studying
Focus on concepts, interpretation, and choosing the right tool for the job. You do not need to memorize long stretches of code.
:::


## What You Should Be Ready To Do {.smaller}

:::: {.columns}

::: {.column width="48%"}
**Multiple choice**

- Identify variable types
- Choose an appropriate probability distribution
- Recognize skew and the normal-curve rules
- Pick the right figure or test
- Distinguish population from sample
- Interpret a p-value or `R^2`
:::

::: {.column width="52%"}
**Short answer**

- Read a t-test/chi-square test
- Interpret regression output
- Read and interpret interaction terms and margins plots
- Explain what moderation means and interpret an interaction coefficient
- Evaluate sampling designs and representativeness
- Critique a misleading figure
- Interpret a confidence interval
:::

::::



## Examples of Choosing the Right Tool {.smaller}

| Goal | What the variables look like | Good choice |
|---|---|---|
| Show the distribution of one categorical variable | One categorical variable | Bar chart |
| Show the distribution of one quantitative variable | One continuous/count variable | Histogram |
| Compare the mean of a quantitative outcome across two groups | Quantitative `Y` + binary group | Two-sample t-test |
| Study the relationship between two categorical variables | Categorical `X` + categorical `Y` | Chi-square test |
| Model a continuous outcome | Quantitative `Y` | OLS regression |
| Model a binary outcome | `Y = 0/1` | Logistic regression |

::: {.callout-note icon=false}
## On the exam
Many mistakes come from choosing the wrong tool because the student did not first identify the outcome type.
:::


## Inference Refresher {.smaller}

- The **68-95-99.7 rule** only applies to an approximately normal distribution
- The **Central Limit Theorem** is about the sampling distribution of sample means, not about every raw sample looking normal
- A **margin of error** is the plus-or-minus amount around an estimate
- A **confidence interval** is the full range: estimate plus or minus the margin of error
- Holding everything else constant, a **larger sample size** usually makes a confidence interval narrower
- A **higher confidence level** usually makes a confidence interval wider

::: {.callout-tip icon=false}
## P-value wording to know well
A p-value is the probability of getting results this extreme, or more extreme, **if the null hypothesis were true**.

For any output you read on the exam: identify the **direction**, comment on **statistical significance**, then interpret **substantively** in plain language.
:::


## Reading OLS Output {.smaller}

| Term | Estimate |
|---|---:|
| Intercept | 5.8*** |
| Divorced | -0.9** |
| Widowed | -0.6 |
| Never married | -1.2*** |
| Age | 0.02* |
| `R^2` | 0.16 |

*Outcome: life satisfaction (1–7 scale). Married is the omitted baseline.*

- In OLS, the outcome is **continuous**
- If one category is missing from the table, that missing category is the **baseline/reference group**
- A coefficient tells you the expected difference in the outcome, **holding the other variables constant**
- `R^2` tells you how much of the variation in the outcome is explained by the model

::: {.callout-tip icon=false}
## Study move
Practice saying coefficient interpretations out loud using the phrase "holding the other variables constant."
:::

::: {.callout-note icon=false}
## Sample interpretation — Never married
Respondents who have never been married are predicted to score 1.2 points lower on life satisfaction than married respondents (the reference group), holding age constant.
:::


## Reading Logistic Regression Output {.smaller}

| Term | Odds ratio |
|---|---:|
| Income (thousands) | 1.02** |
| Age | 1.03*** |
| Married (1 = yes) | 1.38* |

*Outcome: whether the respondent participates in a community organization (1 = yes).*

- In logistic regression, the outcome is **binary**
- Odds ratio greater than 1 -> **higher odds**
- Odds ratio less than 1 -> **lower odds**
- Odds ratio equal to 1 -> **no association**
- A quick interpretation shortcut is `(OR - 1) x 100`

**Examples**

- `1.38` -> about **38% higher odds**
- `0.76` -> about **24% lower odds**

::: {.callout-note icon=false}
## Sample interpretation — Age
Each additional year of age is associated with odds of participating in a community organization that are about 3% higher, holding income and marital status constant.
:::


## Interactions and Margins Plots {.smaller}

An interaction asks whether the relationship between `X` and `Y` depends on a third variable.

```{r}
#| echo: false
#| fig-height: 4.2
#| fig-align: center
print(interaction_plot)
```

- **Parallel lines** suggest little or no moderation
- **Non-parallel lines** suggest the slope changes across groups or levels
- The lower-order terms and the interaction term all matter for interpretation

::: {.callout-note icon=false}
## Sample interpretation
Group A's education slope (1.2) is steeper than Group B's (0.4) — group membership moderates the relationship between education and the outcome.
:::

## Moderation {.smaller}

| How to Identify | What Moderation does |
|-|---|
| Big question | **For whom** or **under what conditions** does `X` affect `Y`? |
| Third variable | Changes the slope |
| Typical model | Interaction term |
| Visual idea | Non-parallel lines |
| Example | Education predicts income differently for men and women |

::: {.callout-tip icon=false}
## Quick test
If a third variable **changes the slope** of `X -> Y` across groups or levels, that is moderation. Look for non-parallel lines in a margins plot.
:::


## Sampling and Figure Critique {.smaller}

:::: {.columns}

::: {.column width="48%"}
**Sampling**

- A representative sample resembles the target population on the characteristics that matter
- In cluster sampling you randomly select groups or locations first, then observe people within them (e.g., randomly select hospitals, then survey all patients inside)
- Nonresponse or selection bias happens when some kinds of people are systematically harder to reach than others
:::

::: {.column width="52%"}
**Figure critique**

- Does the graph type match the variable?
- Are the axes labeled clearly?
- Is the title accurate?
- Is the scale misleading or truncated in a way that exaggerates differences?
- Are the units clear?
:::

::::


## How to Study Efficiently This Week {.smaller}

- Revisit the slides and assignments on these topics:
  - Weeks 7-9: sampling, distributions, CLT, confidence intervals, hypothesis tests
  - Week 11: OLS regression and `R^2`
  - Week 12: logistic regression and odds ratios
  - Week 13: interactions (moderation) and margins plots
- Redo one example each of a t-test, an OLS model, a logistic model, and an interaction model
- Practice turning output into short written interpretations
- Review definitions you want to be able to write cleanly from memory:
  - p-value
  - confidence interval
  - baseline/reference group
  - moderation
  - representative sample

::: {.callout-note icon=false}
## Best use of your time
Do not just reread code. Practice explaining what results mean.
:::


## Questions?

- What topics still feel fuzzy?
- Which type of output do you want to practice interpreting one more time?
