---
title: "Week 14"
subtitle: "Sociology 106: Quantitative Sociological Methods"
date: 04/21/2026
date-format: long
format:
  revealjs:
    slide-number: true
    chalkboard:
      buttons: false
    preview-links: auto
    css: slides.css
    scrollable: true
    footer: '[GitHub course repo](https://www.kaseyzapatka.com/soc106)'
---

```{r}
#| echo: false
#| results: false
#| warning: false
#| message: false

# will likely need to install
# install.packages("mediation")

library(tidyverse)
library(here)
library(patchwork)
library(modelsummary)
library(ggeffects)
library(mediation)
library(car)        # for vif()

options(scipen = 999)
options(modelsummary_factory_html = "gt")

attain <- read_csv(here("data/attain.csv"))

set.seed(123)
attain <- attain |>
  mutate(
    prestg80 = pmax(15, pmin(97, prestg80 + rnorm(n(), mean = 0, sd = 10))),
    income91 = pmax(5000, income91 * exp(rnorm(n(), mean = 0, sd = 0.45)))
  )

attain <- attain |>
  mutate(
    degree_f     = fct_relevel(degree, "lt high", "high sch", "junior c", "bachelor", "graduate"),
    sex_f        = fct_relevel(sex, "male"),
    union_member = as.integer(union %in% c("r belong", "r and sp"))
  )

# Complete-case subset for mediation analysis
# Key path: Education (X) → Occupational Prestige (M) → Income (Y)
attain_med <- attain |>
  filter(!is.na(income91), !is.na(educ), !is.na(prestg80), !is.na(sex_f))

# ---- MEDIATION MODELS ----
# Step 1 (total effect): educ → income
med_total   <- lm(income91  ~ educ,                    data = attain_med)
# Step 2 (X → M):        educ → prestg80
med_a_path  <- lm(prestg80  ~ educ,                    data = attain_med)
# Step 3 (X + M → Y):    educ + prestg80 → income
med_b_path  <- lm(income91  ~ educ + prestg80,         data = attain_med)

# Coefficients for inline text
c_prime      <- round(coef(med_total)[["educ"]])          # total effect
a_coef       <- round(coef(med_a_path)[["educ"]], 2)      # a path
b_coef       <- round(coef(med_b_path)[["prestg80"]])     # b path
c_direct     <- round(coef(med_b_path)[["educ"]])         # direct effect
indirect_est <- round(a_coef * b_coef)                    # approx indirect

# Mediation package: bootstrap CIs for indirect effect
med_out <- mediate(
  model.m = med_a_path,
  model.y = med_b_path,
  treat   = "educ",
  mediator = "prestg80",
  boot    = TRUE,
  sims    = 500
)

acme_est  <- round(med_out$d0)
acme_lo   <- round(med_out$d0.ci[[1]])
acme_hi   <- round(med_out$d0.ci[[2]])
ade_est   <- round(med_out$z0)
total_est <- round(med_out$tau.coef)

# ---- MULTICOLLINEARITY MODELS ----
# High collinearity: educ + degree_f (very correlated)
model_vif_high <- lm(income91 ~ educ + prestg80 + age, data = attain_med)
model_vif_bad  <- lm(income91 ~ educ + prestg80 + age +
                       as.numeric(degree_f),            data = attain_med)
vif_vals <- round(vif(model_vif_high), 2)

# Pre-compute for slide text
pct_mediated <- round(100 * abs(acme_est) / abs(total_est))
```


## Agenda {.smaller}

**Housekeeping:**

- Final paper, HW #11 (canceled), In-class presentations

**Statistical content — three parts:**

- **Part 1**: Mediation vs. moderation — getting the concepts straight
- **Part 2**: Mediation analysis — path models, Baron & Kenny, R implementation
- **Part 3**: Multicollinearity — when predictors are too similar


## Housekeeping {.smaller}

**Weekly Assignment #11**

- HW 11 was canceled - wanted to make sure you had time to work on final exam 
- Will still be able to drop lowest of 10 assignments.

**In-class presentations**

- <span class="highlight">Tuesday, April 28/Tuesday May, 5</span> — 7–10 minutes per student
- Slide template on the course site; structure: question → data → results → so what
- Practice out loud at least once before class

**Final paper**

- <span class="highlight">Due Thursday, May 7</span>
- Should include a regression model that answers your research question
- Should follow the formatting described in the [final paper outline](https://www.kaseyzapatka.com/soc106/assignments/paper/03_proposal_outline.html) 


## Where We Are in the Course {.smaller}

- **Week 11:** OLS regression — estimating linear relationships
- **Week 12:** Logistic regression — predicting binary outcomes
- **Week 13:** Interaction effects — *for whom* does X affect Y?
- **This week (Week 14):** Mediation — *how* or *why* does X affect Y?

::: {.callout-note icon=false}
## The arc of the course: from "Is there a relationship?" to "What kind?"
Each week we've added a layer of nuance. This week we ask not just whether X predicts Y, but **through what mechanism** — and we tackle a practical issue (multicollinearity) you'll likely encounter writing up your papers.
:::


## Questions?


# Part 1: Getting the Concepts Straight

Moderation vs. Mediation — the distinction that trips everyone up


## Two Questions, Two Models {.smaller}

::: {.panel-tabset}

### Side-by-Side

| | **Moderation** | **Mediation** |
|---|---|---|
| **Question** | *For whom* does X affect Y? | *How/why* does X affect Y? |
| **Third variable** | A moderator — *changes the slope* of X→Y | A mediator — *carries the effect* of X to Y |
| **Path structure** | X → Y, with slope depending on Z | X → M → Y |
| **Regression term** | Interaction: `X * Z` | Sequential models: `Y ~ X`, then `Y ~ X + M` |
| **What you see** | Different slopes for different groups | X effect shrinks when M is added |

### Education & Income: Two Different Stories

:::: {.columns}

::: {.column width="50%"}
**Mediation story:**

Does education raise income *because* it leads to higher-prestige jobs?

$$\text{Education} \rightarrow \underbrace{\text{Job Prestige}}_{\text{mediator } M} \rightarrow \text{Income}$$

The mediator *explains the mechanism*.
:::

::: {.column width="50%"}
**Moderation story (Week 13):**

Does education raise income *more* for men than for women?

$$\text{Education} \xrightarrow{\text{slope differs by}} \text{Income}$$
$$\text{(depending on Sex)}$$

The moderator *changes who benefits*.
:::

::::

::: {.callout-tip icon=false}
## Memory trick
**Mediation** → the mechanism (**M**ediator = **M**echanism).
**Moderation** → the modifier (**M**oderator = **M**odifier of the slope).
:::

### Common Mistakes

::: {.fragment .fade-up}
**Mistake 1: Controlling for a mediator**

If you add job prestige as a *control* in a model predicting income from education, you **block the very path you're trying to study**. The coefficient on education will tell you only the direct effect — missing the part that works *through* prestige.

→ Week 13 warned: don't control for mediators. Today we learn what to do instead.
:::

::: {.fragment .fade-up}
**Mistake 2: Calling a mediator a moderator**

A mediator is *on the causal path*. A moderator is *off the path*, conditioning how strong it is.

- "Prestige carries the education→income effect" → **mediator**
- "Sex changes how strong the education→income effect is" → **moderator**
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## The test: where is the third variable?
Draw the arrows. If the third variable sits **between** X and Y (X → M → Y), it's a mediator. If it sits **beside** the X→Y arrow, changing its slope, it's a moderator.
:::
:::

:::


# Part 2: Mediation Analysis

How and why does X affect Y?


## The Three-Effect Framework {.smaller}

```{r}
#| echo: false
#| fig-height: 4.0
#| fig-align: center
#| warning: false

# ---- Path diagram: total vs mediated ----
# Two-panel: left = total effect, right = decomposed

make_path_plot <- function(show_m = FALSE, title_text = "") {
  p <- ggplot() +
    # X node
    annotate("label", x = 0, y = 0,
             label = "Education\n(X)",
             fill = "#d0e4f5", color = "#4E79A7",
             size = 4.5, fontface = "bold", label.size = 1.0) +
    # Y node
    annotate("label", x = 4, y = 0,
             label = "Income\n(Y)",
             fill = "#d0e4f5", color = "#4E79A7",
             size = 4.5, fontface = "bold", label.size = 1.0) +
    labs(title = title_text) +
    coord_cartesian(xlim = c(-0.7, 4.7), ylim = c(-0.7, 2.5)) +
    theme_void(base_size = 13) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13))

  if (!show_m) {
    # Total effect: single arrow X → Y
    p <- p +
      annotate("segment",
               x = 0.52, xend = 3.48, y = 0, yend = 0,
               arrow = arrow(type = "closed", length = unit(0.15, "in")),
               color = "#4E79A7", linewidth = 1.6) +
      annotate("text", x = 2, y = 0.28,
               label = paste0("c' (total = $", scales::comma(total_est), "/yr)"),
               size = 3.8, color = "#4E79A7", fontface = "bold")
  } else {
    # Mediator node
    p <- p +
      annotate("label", x = 2, y = 2.0,
               label = "Job Prestige\n(M)",
               fill = "#e8f5e9", color = "#59A14F",
               size = 4.5, fontface = "bold", label.size = 1.0) +
      # a path: X → M
      annotate("segment",
               x = 0.52, xend = 1.55, y = 0.28, yend = 1.74,
               arrow = arrow(type = "closed", length = unit(0.13, "in")),
               color = "#59A14F", linewidth = 1.4) +
      annotate("text", x = 0.72, y = 1.15,
               label = paste0("a = ", a_coef),
               size = 3.8, color = "#59A14F", fontface = "bold") +
      # b path: M → Y
      annotate("segment",
               x = 2.45, xend = 3.48, y = 1.74, yend = 0.28,
               arrow = arrow(type = "closed", length = unit(0.13, "in")),
               color = "#59A14F", linewidth = 1.4) +
      annotate("text", x = 3.28, y = 1.15,
               label = paste0("b = $", scales::comma(b_coef)),
               size = 3.8, color = "#59A14F", fontface = "bold") +
      # Direct effect: X → Y
      annotate("segment",
               x = 0.52, xend = 3.48, y = -0.05, yend = -0.05,
               arrow = arrow(type = "closed", length = unit(0.13, "in")),
               color = "#F28E2B", linewidth = 1.4) +
      annotate("text", x = 2, y = -0.38,
               label = paste0("c (direct = $", scales::comma(c_direct), "/yr)"),
               size = 3.8, color = "#F28E2B", fontface = "bold")
  }
  p
}

p_total <- make_path_plot(show_m = FALSE, title_text = "Total Effect")
p_decomp <- make_path_plot(show_m = TRUE,  title_text = "Decomposed")

p_total + p_decomp
```

::: {.fragment .fade-up}
$$\underbrace{c'}_{\text{total}} = \underbrace{c}_{\text{direct}} + \underbrace{a \times b}_{\text{indirect}}$$

- **Total effect** (*c'*): the overall X → Y relationship (before adding M)
- **Direct effect** (*c*): X → Y *after* controlling for M
- **Indirect effect** (*a × b*): the part of X's effect that travels *through* M
:::


## The *a*, *b*, *c* Paths Explained {.smaller}

:::: {.columns}

::: {.column width="55%"}

::: {.panel-tabset}

### Path *a*: X → M

**Does education predict job prestige?**

```{r}
#| echo: true
#| eval: false
model_a <- lm(prestg80 ~ educ, data = attain_med)
```

```{r}
#| echo: false
modelsummary(med_a_path,
             coef_rename = c("educ" = "Education (yrs)"),
             statistic = "({std.error})",
             gof_map = c("nobs", "r.squared"),
             fmt = 2,
             title = "Path a: Education → Prestige")
```

Each additional year of education is associated with **`r a_coef` more prestige points**.

### Path *b*: M → Y (+ X)

**Does prestige predict income, controlling for education?**

```{r}
#| echo: true
#| eval: false
model_b <- lm(income91 ~ educ + prestg80, data = attain_med)
```

```{r}
#| echo: false
modelsummary(med_b_path,
             coef_rename = c("educ" = "Education (yrs)",
                             "prestg80" = "Job Prestige"),
             statistic = "({std.error})",
             gof_map = c("nobs", "r.squared"),
             fmt = 0,
             title = "Paths b & c: Educ + Prestige → Income")
```

### Total Effect (*c'*)

**Ignoring prestige — what's the raw education effect?**

```{r}
#| echo: true
#| eval: false
model_total <- lm(income91 ~ educ, data = attain_med)
```

```{r}
#| echo: false
modelsummary(med_total,
             coef_rename = c("educ" = "Education (yrs)"),
             statistic = "({std.error})",
             gof_map = c("nobs", "r.squared"),
             fmt = 0,
             title = "Total Effect: Education → Income")
```

Total effect: **$`r scales::comma(c_prime)`/year of education**.

:::

:::

::: {.column width="45%"}

::: {.callout-note icon=false}
## Putting it together

- **Total effect**: $`r scales::comma(c_prime)` per year of educ
- **Direct effect**: $`r scales::comma(c_direct)` per year of educ
- **Indirect effect** (*a × b*): `r a_coef` × $`r scales::comma(b_coef)` ≈ $`r scales::comma(indirect_est)`

A year of education raises prestige by `r a_coef` points; each prestige point raises income by $`r scales::comma(b_coef)`. So **about `r pct_mediated`% of the education effect on income runs through job prestige**.
:::

::: {.callout-tip icon=false}
## Why not just subtract?
You can estimate the indirect effect as *c' − c*, but the *a × b* product gives you the same number and generalizes better (especially with bootstrapped confidence intervals).
:::

:::

::::


## Full vs. Partial Mediation {.smaller}

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-align: center
#| warning: false

# Illustrate full vs partial mediation with simulated data
set.seed(42)
n_ex <- 400
x_ex  <- rnorm(n_ex, 12, 3)
m_ex  <- 0.8 * x_ex + rnorm(n_ex, 0, 2)

# Full mediation: direct effect ≈ 0 after controlling for M
y_full  <- 5000 + 1200 * m_ex              + rnorm(n_ex, 0, 3000)
# Partial mediation: some direct effect remains
y_part  <- 5000 + 1200 * m_ex + 1500 * x_ex + rnorm(n_ex, 0, 3000)

df_full <- tibble(x = x_ex, m = m_ex, y = y_full, type = "Full mediation")
df_part <- tibble(x = x_ex, m = m_ex, y = y_part, type = "Partial mediation")

df_plot <- bind_rows(df_full, df_part)

ggplot(df_plot, aes(x = x, y = y)) +
  geom_point(alpha = 0.18, size = 0.9, color = "#4E79A7") +
  geom_smooth(method = "lm", se = FALSE, color = "#E15759", linewidth = 1.1) +
  geom_smooth(data = df_plot, aes(x = m, y = y),
              method = "lm", se = FALSE, color = "#59A14F",
              linewidth = 1.1, linetype = "dashed") +
  facet_wrap(~type) +
  labs(
    x = "X (or M — green dashed)",
    y = "Y (Income)",
    caption = "Red = X→Y; Green dashed = M→Y after controlling for X"
  ) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold", size = 13))
```

:::: {.columns}

::: {.column width="50%"}
**Full mediation**

The direct effect *c* drops to (near) **zero** when M enters the model.

M fully explains *why* X affects Y.

*The indirect path is the whole story.*
:::

::: {.column width="50%"}
**Partial mediation**

The direct effect *c* decreases but **remains significant**.

M explains *part* of the mechanism; something else also connects X directly to Y.

*Both paths matter.*
:::

::::

::: {.callout-tip icon=false}
## In practice: partial mediation is the norm
Full mediation is rare in observational social science. Expect the direct effect to shrink — but not disappear. That's still an important substantive finding.
:::


## Baron & Kenny Four Steps {.smaller .scrollable}

::: {.panel-tabset}

### The Steps

The Baron & Kenny (1986) approach tests mediation through four sequential regressions:

::: {.fragment .fade-up}
| Step | Model | What you're checking |
|---|---|---|
| **1** | `lm(Y ~ X)` | X significantly predicts Y (there's something to explain) |
| **2** | `lm(M ~ X)` | X significantly predicts M (X moves the mediator) |
| **3** | `lm(Y ~ X + M)` | M significantly predicts Y, controlling for X |
| **4** | Compare Step 1 vs. Step 3 | X→Y coefficient *decreases* when M is added |
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## The logic
If X moves M (Step 2), and M moves Y independently of X (Step 3), then some of what looks like a direct X→Y effect is really X working *through* M. Step 4 quantifies how much.
:::
:::

### Step-by-Step: Our Example

```{r}
#| echo: true

# Step 1: Does education predict income?
step1 <- lm(income91 ~ educ,            data = attain_med)

# Step 2: Does education predict prestige?
step2 <- lm(prestg80 ~ educ,            data = attain_med)

# Step 3 & 4: Does prestige predict income? Does education effect shrink?
step3 <- lm(income91 ~ educ + prestg80, data = attain_med)
```

```{r}
#| echo: false
modelsummary(
  list("Step 1: Y~X" = step1,
       "Step 2: M~X" = step2,
       "Step 3: Y~X+M" = step3),
  coef_rename = c("educ"     = "Education (X)",
                  "prestg80" = "Job Prestige (M)"),
  statistic = "({std.error})",
  stars = TRUE,
  gof_map = c("nobs", "r.squared"),
  fmt = 0,
  title = "Baron & Kenny Steps 1–4"
)
```

### Interpreting the Table

**What the table shows:**

- **Step 1**: Education → Income: +$`r scales::comma(c_prime)` per year ✓ (Step 1 met)
- **Step 2**: Education → Prestige: +`r a_coef` pts per year ✓ (Step 2 met)
- **Step 3**: Prestige → Income (controlling educ): +$`r scales::comma(b_coef)` per prestige point ✓ (Step 3 met)
- **Step 4**: Education effect on income drops from $`r scales::comma(c_prime)` → $`r scales::comma(c_direct)` when prestige enters ✓

::: {.callout-tip icon=false}
## Verdict: **partial mediation**
Education still has a direct effect on income, but part of its total effect works *through* occupational prestige. Both the direct and indirect paths matter.
:::

### Limitation of B&K

Baron & Kenny tells you *whether* mediation exists, but not *how much* or *how confident* you should be in the indirect effect.

::: {.fragment .fade-up}
The indirect effect (*a × b*) needs its **own standard error and confidence interval** — and the distribution of a product of two estimates is not normal, so we can't use a simple z-test.

**Solution: bootstrapping** (re-sample the data many times, compute *a × b* each time, use the percentiles as the CI).
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## Modern practice
The `mediation` package in R implements bootstrapped confidence intervals for the indirect effect (ACME = Average Causal Mediation Effect). This is the standard in published sociology papers.
:::
:::

:::


## Indirect Effects with `mediation` {.smaller}

:::: {.columns}

::: {.column width="55%"}

```{r}
#| echo: true
#| eval: false

library(mediation)

# Fit the two models first
model_m <- lm(prestg80 ~ educ,            data = attain_med)
model_y <- lm(income91 ~ educ + prestg80, data = attain_med)

# Run mediation with bootstrap
med_out <- mediate(
  model.m  = model_m,
  model.y  = model_y,
  treat    = "educ",       # X variable
  mediator = "prestg80",   # M variable
  boot     = TRUE,
  sims     = 1000          # bootstrap draws
)

summary(med_out)
```

::: {.callout-note icon=false}
## Reading the output

- **ACME** = Average Causal Mediation Effect = indirect effect (*a × b*)
- **ADE** = Average Direct Effect = direct effect (*c*)
- **Total Effect** = ACME + ADE (*c'*)
- **Prop. Mediated** = ACME / Total Effect
:::

:::

::: {.column width="45%"}

**Our results (500 bootstraps):**

| Effect | Estimate | 95% CI |
|---|---|---|
| **Indirect (ACME)** | $`r scales::comma(acme_est)` | [$`r scales::comma(acme_lo)`, $`r scales::comma(acme_hi)`] |
| **Direct (ADE)** | $`r scales::comma(ade_est)` | — |
| **Total** | $`r scales::comma(total_est)` | — |
| **% Mediated** | `r pct_mediated`% | — |

::: {.fragment .fade-up}
::: {.callout-tip icon=false}
## Interpretation
About **`r pct_mediated`% of education's effect on income** is mediated by occupational prestige. The indirect path (education → prestige → income) is statistically distinguishable from zero (the CI excludes 0). But a substantial direct effect remains.
:::
:::

:::

::::


## Visualizing the Indirect Effect {.smaller}

```{r}
#| echo: false
#| fig-height: 4.0
#| fig-align: center
#| warning: false

# Coefficient plot showing how education coefficient changes
coef_data <- tibble(
  model   = c("Total effect\n(no mediator)", "Direct effect\n(with prestige)"),
  estimate = c(c_prime, c_direct),
  se      = c(
    summary(med_total)$coefficients["educ", "Std. Error"],
    summary(med_b_path)$coefficients["educ", "Std. Error"]
  )
) |>
  mutate(
    lo = estimate - 1.96 * se,
    hi = estimate + 1.96 * se,
    model = fct_rev(fct_inorder(model))
  )

ggplot(coef_data, aes(x = estimate, y = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60", linewidth = 0.8) +
  geom_segment(aes(x = lo, xend = hi, yend = model),
               linewidth = 2.2, color = "#4E79A7", alpha = 0.4) +
  geom_point(size = 5, color = "#4E79A7") +
  geom_text(aes(label = paste0("$", scales::comma(round(estimate)))),
            hjust = -0.25, size = 4.2, fontface = "bold", color = "#4E79A7") +
  annotate("segment",
           x = c_direct + 120, xend = c_prime - 120,
           y = 1.5, yend = 1.5,
           arrow = arrow(ends = "both", type = "closed",
                         length = unit(0.12, "in")),
           color = "#59A14F", linewidth = 1.3) +
  annotate("text",
           x = (c_direct + c_prime) / 2, y = 1.65,
           label = paste0("Indirect\n$", scales::comma(indirect_est)),
           size = 3.8, color = "#59A14F", fontface = "bold") +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Effect of one additional year of education on income",
    y = NULL,
    title = "Mediation shrinks the education coefficient",
    subtitle = "Adding job prestige absorbs part of the education effect"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.y  = element_text(size = 12),
    plot.title   = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40")
  )
```


## Causal Limitations {.smaller}

::: {.panel-tabset}

### What Mediation Can't Do

Mediation analysis tests a **causal claim** — that X works *through* M — but with observational data, we can never fully prove it.

::: {.fragment .fade-up}
**Three threats to causal mediation:**

| Threat | What it means |
|---|---|
| **No temporal ordering** | We assume X precedes M precedes Y. With cross-sectional data, we can't verify this. |
| **M→Y confounding** | A third variable may cause both prestige and income, creating a spurious M→Y association. |
| **X→M→Y vs X←M→Y** | Without experiments, the arrow could point the other way (prestige affects education attainment). |
:::

::: {.fragment .fade-up}
::: {.callout-tip icon=false}
## The gold standard
The cleanest mediation evidence comes from **randomized experiments** where X is randomly assigned. Then you know X precedes M, and there's no X-confounder. But experiments are rare in sociology — so we work with what we have, acknowledge the limits, and make the theoretical argument carefully.
:::
:::

### For Your Final Papers

**If you're running a mediation analysis:**

1. Fit the three models (B&K Steps 1–3)
2. Report coefficients for all three — show the reader the direct, indirect, and total effects
3. Use `mediate()` with `boot = TRUE` for the indirect effect CI
4. State your theoretical argument for *why* M mediates X→Y
5. Acknowledge the cross-sectional limitation — you can't prove causal mediation, but you can show the pattern is consistent with it

::: {.fragment .fade-up}
**Language template:**

> "Consistent with a mediation hypothesis, the coefficient on [X] declined from [total] to [direct] when [M] was included (ACME = [value], 95% CI [lo, hi]). This suggests that approximately [%] of the association between [X] and [Y] may operate through [M], though cross-sectional data preclude strong causal claims about the mediating mechanism."
:::

### What to Do Instead (Advanced)

If you wanted stronger causal mediation evidence with observational data, you would:

- Use **panel data** (before/after measurements) to verify temporal ordering
- Control for all plausible **M→Y confounders** explicitly
- Use **sensitivity analysis** (`medsens()` in the `mediation` package) to test how robust the indirect effect is to unmeasured M→Y confounding

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## For this course
For your final papers: fit the three models, bootstrap the indirect effect, state the caveats. That's the sociological standard for cross-sectional mediation claims.
:::
:::

:::


# Part 3: Multicollinearity

When your predictors are too alike


## What Is Multicollinearity? {.smaller}

**Multicollinearity** occurs when two or more predictors in a regression are highly correlated with each other.

:::: {.columns}

::: {.column width="50%"}

::: {.fragment .fade-up}
**Why it's a problem:**

When X₁ and X₂ are highly correlated:
- The model can't tell which one is doing the work
- Coefficient estimates become **unstable** — tiny data changes → large coefficient changes
- **Standard errors inflate** → wide confidence intervals → everything looks non-significant
- Coefficients can flip sign or become implausibly large
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## The core issue
Multicollinearity doesn't bias your coefficients — it just makes them imprecise. With enough data, standard errors shrink. But in typical survey samples, high multicollinearity can completely obscure real effects.
:::
:::

:::

::: {.column width="50%"}

::: {.fragment .fade-up}
```{r}
#| echo: false
#| fig-height: 4.0
#| fig-align: center
#| warning: false

# Illustrate: educ and years of prestige are correlated
cor_val <- round(cor(attain_med$educ, attain_med$prestg80, use = "complete.obs"), 2)

ggplot(attain_med, aes(x = educ, y = prestg80)) +
  geom_point(alpha = 0.25, size = 0.9, color = "#4E79A7") +
  geom_smooth(method = "lm", se = TRUE, color = "#E15759",
              fill = "#E15759", alpha = 0.15, linewidth = 1.2) +
  labs(
    x = "Education (years)",
    y = "Occupational Prestige",
    title = paste0("Education and prestige: r = ", cor_val),
    subtitle = "Correlated — but not perfectly so (both can be in the model)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```
:::

:::

::::


## Detecting Multicollinearity: VIF {.smaller}

:::: {.columns}

::: {.column width="55%"}

The **Variance Inflation Factor (VIF)** measures how much each predictor's variance is inflated by correlation with the others.

$$\text{VIF}_j = \frac{1}{1 - R^2_j}$$

where $R^2_j$ is the R² from regressing predictor $j$ on all other predictors.

::: {.fragment .fade-up}
**Rules of thumb:**

| VIF | Interpretation |
|---|---|
| 1 | No multicollinearity |
| 1–5 | Low — acceptable |
| 5–10 | Moderate — investigate |
| > 10 | Severe — take action |
:::

:::

::: {.column width="45%"}

```{r}
#| echo: true

library(car)

# Fit model with education, prestige, and age
model_check <- lm(
  income91 ~ educ + prestg80 + age,
  data = attain_med
)

vif(model_check)
```

```{r}
#| echo: false
#| fig-height: 2.8
#| warning: false

vif_df <- tibble(
  predictor = names(vif_vals),
  vif       = vif_vals
) |>
  mutate(predictor = c("Education", "Job Prestige", "Age"),
         color = case_when(vif > 10 ~ "#E15759",
                           vif > 5  ~ "#F28E2B",
                           TRUE     ~ "#59A14F"))

ggplot(vif_df, aes(x = vif, y = fct_reorder(predictor, vif), fill = color)) +
  geom_col(width = 0.5) +
  geom_vline(xintercept = 5,  linetype = "dashed", color = "#F28E2B", linewidth = 0.8) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "#E15759", linewidth = 0.8) +
  geom_text(aes(label = vif), hjust = -0.3, size = 4) +
  scale_fill_identity() +
  scale_x_continuous(limits = c(0, max(vif_df$vif) * 1.3)) +
  labs(x = "VIF", y = NULL,
       title = "VIF values — all well below 5") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

:::

::::


## What Multicollinearity Looks Like {.smaller}

```{r}
#| echo: false
#| fig-height: 4.2
#| fig-align: center
#| warning: false

# Manually specified for clarity: identical point estimates, SEs inflated by
# VIF = 1/(1 - r²) under high multicollinearity (r = 0.95 → VIF ≈ 10.3 → SE × 3.2)
sim_df <- tribble(
  ~condition,                          ~term, ~estimate, ~se,
  "Low multicollinearity (r ≈ 0)",     "X₁",       2.0, 0.09,
  "Low multicollinearity (r ≈ 0)",     "X₂",       1.5, 0.09,
  "High multicollinearity (r ≈ 0.95)", "X₁",       2.0, 0.29,
  "High multicollinearity (r ≈ 0.95)", "X₂",       1.5, 0.29
) |>
  mutate(lo        = estimate - 1.96 * se,
         hi        = estimate + 1.96 * se,
         condition = fct_inorder(condition))

true_vals <- tibble(term = c("X₁", "X₂"), truth = c(2, 1.5))

ggplot(sim_df, aes(x = term, y = estimate, color = condition)) +
  geom_hline(data = true_vals, aes(yintercept = truth),
             linetype = "dotted", color = "gray40", linewidth = 0.8) +
  geom_errorbar(aes(ymin = lo, ymax = hi),
                width = 0.15, linewidth = 1.3,
                position = position_dodge(width = 0.4)) +
  geom_point(size = 4.5, position = position_dodge(width = 0.4)) +
  scale_color_manual(values = c("#59A14F", "#E15759"), name = NULL) +
  annotate("text", x = 0.62, y = 2.08,
           label = "true = 2",   size = 3.5, color = "gray40", fontface = "italic") +
  annotate("text", x = 1.62, y = 1.58,
           label = "true = 1.5", size = 3.5, color = "gray40", fontface = "italic") +
  labs(
    x        = NULL,
    y        = "Coefficient estimate (95% CI)",
    title    = "Multicollinearity inflates standard errors — not the estimates",
    subtitle = "Identical point estimates; SEs are ~3× larger when r(X₁, X₂) = 0.95"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position  = "bottom",
    plot.title       = element_text(face = "bold"),
    plot.subtitle    = element_text(color = "gray40")
  )
```

::: {.callout-tip icon=false}
## The takeaway
The point estimates are approximately the same in both conditions — multicollinearity does **not** bias coefficients. What it does is inflate the standard errors: the confidence intervals are much wider when x₁ and x₂ are highly correlated. You're more likely to miss a real effect (Type II error), but the estimates themselves are not systematically wrong.
:::


## Is Multicollinearity Actually a Problem? {.smaller}

::: {.panel-tabset}

### The Short Answer

**It depends on what you're trying to do — and which variables are collinear.**

| Goal | Does multicollinearity matter? |
|---|---|
| **Prediction** | Usually **no** — collinear models predict just as well; R² is unaffected |
| **Inference on a control variable** | Often **no** — imprecise control coefficients don't invalidate your key result |
| **Inference on your variable of interest** | **Yes** — if your key X is the collinear one, SEs inflate and you may miss a real effect |
| **Understanding relative importance of two predictors** | **Yes** — you can't disentangle their individual contributions |

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## The core issue: precision, not accuracy
Multicollinearity does **not bias your coefficients** — the estimates are still correct on average. What it does is make them *imprecise*: standard errors inflate, confidence intervals widen, and t-statistics shrink. The main consequence is **more Type II errors** — failing to detect real effects that are actually there.
:::
:::

### Type II Errors

**Type II error** = failing to reject H₀ when it's actually false (a "false negative").

::: {.fragment .fade-up}
Here's the chain of consequences:

$$\underbrace{\text{High multicollinearity}}_{\text{predictors too similar}} \Rightarrow \underbrace{\uparrow \text{SE}}_{\text{wider uncertainty}} \Rightarrow \underbrace{\downarrow t\text{-statistic}}_{\frac{\hat\beta}{SE}} \Rightarrow \underbrace{\uparrow p\text{-value}}_{\text{harder to reject }H_0} \Rightarrow \underbrace{\text{Type II error}}_{\text{miss a real effect}}$$
:::

::: {.fragment .fade-up}
**Example:**

Suppose education truly does affect income controlling for prestige (direct effect = $800/year). With low multicollinearity between educ and prestige, SE = 200, t = 4.0, p < 0.001. With high multicollinearity (r = 0.92), SE = 1100, t = 0.73, p = 0.47 — we'd conclude "no direct effect" even though one exists.

The effect is real. The data just can't tell the two predictors apart.
:::

::: {.fragment .fade-up}
::: {.callout-tip icon=false}
## The practical implication
If you run a multivariate model and a variable you expected to be significant is not — and its VIF is high — consider whether multicollinearity is masking the effect. This is *not* grounds to drop the variable; it's grounds to note the limitation.
:::
:::

### It's Not Always Pairwise

Multicollinearity is not just about two variables being correlated. **Any set of predictors can jointly cause it**, even if no two are highly correlated in isolation.

::: {.fragment .fade-up}
**Example 1: Age, work experience, and tenure**

- Age and years of work experience: r ≈ 0.75 (moderate)
- Age and job tenure: r ≈ 0.55 (moderate)
- Experience and tenure: r ≈ 0.60 (moderate)
- But **together**, they carry almost the same information → VIF for each could be 8+

No single pairwise correlation screams "problem," but the three together are nearly redundant.
:::

::: {.fragment .fade-up}
**Example 2: Education, degree, and test score**

- Education in years, highest degree attained, and a cognitive ability test score
- All three measure roughly the same underlying "human capital" concept
- Including all three → VIF > 10 on each, unstable coefficients, possibly sign flips

→ **This is why you check VIF on the fitted model, not just pairwise correlations.**
:::

::: {.fragment .fade-up}
```{r}
#| echo: true
#| eval: false
# Always check VIF on the full model, not just cor()
library(car)
vif(your_model)   # catches joint multicollinearity that cor() misses
```
:::

### When It's Not Your Variable of Interest

**The best news: collinear controls often don't threaten your main conclusion.**

If your research question is *"Does education affect income?"* and education has a low VIF, then:

- The coefficient on education is precisely estimated ✓
- Its SE and p-value are unaffected by multicollinearity among the controls ✓
- You can still make your main inference confidently ✓

The collinear controls (say, age and experience) just have imprecise coefficients — but you probably don't care about their individual effects.

::: {.fragment .fade-up}
**Example from our data:**

```{r}
#| echo: true

# education is our key IV; age and prestige are controls
model_key <- lm(income91 ~ educ + prestg80 + age, data = attain_med)
vif(model_key)
```

If `educ` has VIF ≈ 2 but `prestg80` and `age` have VIF ≈ 4, that's fine — our inference about education is unaffected.
:::

::: {.fragment .fade-up}
::: {.callout-tip icon=false}
## Rule of thumb for your papers
Check VIF on all predictors. If the **variable of interest** has low VIF (< 5), report it and move on. If **only controls** have elevated VIF, note it briefly but don't panic — your main finding is still valid.
:::
:::

:::


## What to Do About Multicollinearity {.smaller}

::: {.panel-tabset}

### Diagnose First

**Step 1: Check the correlation matrix**

```{r}
#| echo: true

attain_med |>
  dplyr::select(educ, prestg80, age) |>
  cor(use = "complete.obs") |>
  round(2)
```

**Step 2: Check VIF after fitting the model**

```{r}
#| echo: true

model_check <- lm(income91 ~ educ + prestg80 + age, data = attain_med)
vif(model_check)
```

If VIF < 5 for all predictors → you're fine. Continue.

### Solutions

| Problem | Solution |
|---|---|
| Including education **and** degree (two measures of same concept) | Drop one; keep the theoretically central one |
| Including age **and** birth year | These are perfectly collinear — pick one |
| Two highly correlated controls | Drop the one less theoretically important |
| Several items measuring the same underlying concept | **Combine into an index** (average or sum the items) |
| Conceptually distinct variables that happen to correlate | Keep both, note the limitation, get more data |

::: {.fragment .fade-up}
**Creating an index when variables measure the same concept:**

When several correlated predictors tap the same underlying idea (e.g., multiple survey items about socioeconomic status, or hours worked across multiple jobs), collapsing them into a single composite reduces multicollinearity and often improves interpretability.

```{r}
#| echo: true
#| eval: false

# Example: respondent works multiple jobs — combine total hours
attain <- attain |>
  mutate(hrs_total = rowMeans(dplyr::select(., hrs1, hrs2), na.rm = TRUE))

# Example: standardise and average two SES-related items into one index
attain <- attain |>
  mutate(
    educ_z    = scale(educ),
    prestg_z  = scale(prestg80),
    ses_index = (educ_z + prestg_z) / 2   # simple equal-weight composite
  )
```

Index variables also carry a **substantive interpretation**: you're saying these items measure one construct, not two separate concepts. Make that theoretical argument in your paper.
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## What NOT to do
Don't use regularization (ridge/LASSO) without understanding it. Don't drop variables that are theoretically important just because of high VIF. Report the issue and argue for your choice.
:::
:::

### For Your Papers

**Common multicollinearity problems in student papers:**

1. **Including both education (years) and degree** → highly correlated (r ≈ 0.80). Pick one.
2. **Including age and years of work experience** → age − 18 ≈ work experience for many. Drop one.
3. **Including income and earnings from multiple sources** → summing them and using total is usually better.

::: {.fragment .fade-up}
**In your write-up:**

```{r}
#| echo: true
#| eval: false

# Always report VIF after your multivariate model
library(car)
vif(your_final_model)
```

A single line: "VIF values ranged from X to Y (all < 5), indicating no problematic multicollinearity."
:::

:::


# Finishing Strong

Getting your final paper across the finish line


## What Your Final Paper Needs {.smaller}

::: {.panel-tabset}

### Model Checklist

**Required components:**

| Component | What to include |
|---|---|
| **Bivariate model** | `lm(Y ~ X)` — your key IV predicting Y |
| **Multivariate model** | Add at least 2 controls; discuss why they matter |
| **Extension** | At least one of: interaction, mediation, or logistic (if binary Y) |
| **Assumption check** | Residual plot + VIF — a paragraph is sufficient |
| **Model table** | `modelsummary()` comparing models side-by-side |

::: {.fragment .fade-up}
```{r}
#| echo: true
#| eval: false

# The pattern for a final paper model table
modelsummary(
  list(
    "Bivariate"     = model_biv,
    "Multivariate"  = model_multi,
    "With Interaction" = model_interact    # or model with mediation
  ),
  stars  = TRUE,
  vcov   = "HC1",           # robust SEs — use these throughout
  gof_map = c("nobs", "r.squared", "adj.r.squared")
)
```
:::

### Interpreting Coefficients

**Don't just report the number — tell the story:**

| Instead of... | Try... |
|---|---|
| "The coefficient on education is 1847." | "Each additional year of education is associated with $1,847 higher income, holding sex and age constant." |
| "The interaction term is significant." | "The education–income association is steeper for men than women: each year of education yields approximately $[gap] more income for men." |
| "ACME = 624, p < 0.05." | "About [%]% of education's effect on income appears to operate through occupational prestige." |

::: {.fragment .fade-up}
::: {.callout-tip icon=false}
## Lead with substance
Sociological audiences want to know: *what does this mean for inequality / stratification / the social world?* Start with the finding, then give the number.
:::
:::

### Writing About Limitations

**Every quantitative paper needs a limitations section. Be specific, not generic.**

Generic (avoid):
> "This study has some limitations. The data may not be perfectly representative."

Specific (better):
> "Because the GSS uses cross-sectional data, we cannot establish the temporal ordering required for causal inference. Specifically, the mediation claim — that education affects income through occupational prestige — requires that education precede prestige, which we assume but cannot verify. Additionally, unmeasured confounders of the prestige–income relationship (e.g., parental occupational networks) may bias the indirect effect estimate."

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## The goal of limitations
Not to undermine your argument — but to show the reader you understand the boundaries of your evidence. A specific limitation paragraph signals methodological maturity.
:::
:::

### Coefficient Plot

A **coefficient plot** shows each predictor's estimated effect and its confidence interval as a dot-and-whisker. It lets readers immediately see which effects are large, which are small, and which cross zero (non-significant).

Use a coefficient plot in your **presentation** instead of the full regression table — tables are hard to read aloud. Use the table in your **paper**.

::: {.fragment .fade-up}
```{r}
#| echo: true
#| fig-height: 3.2
#| fig-align: center
#| warning: false

library(broom)

model_coefplot <- lm(income91 ~ educ + prestg80 + sex_f + age,
                     data = attain_med)

tidy(model_coefplot, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(term = case_match(term,
    "educ"        ~ "Education (yrs)",
    "prestg80"    ~ "Job Prestige",
    "sex_ffemale" ~ "Female",
    "age"         ~ "Age"
  )) |>
  ggplot(aes(x = estimate, y = fct_reorder(term, estimate))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, linewidth = 1.1, color = "#4E79A7") +
  geom_point(size = 3.5, color = "#4E79A7") +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(x = "Effect on income (95% CI)", y = NULL,
       title = "Predictors of income — coefficient plot") +
  theme_minimal(base_size = 12)
```
:::

::: {.fragment .fade-up}
::: {.callout-tip icon=false}
## Reading the plot
A dot to the right of zero means a positive effect; left means negative. If the whisker (CI) crosses the dashed line at zero, the effect is not statistically significant at the 5% level. **The distance from zero shows effect size; the width of the whisker shows precision.**
:::
:::

:::


## Key Takeaways {.smaller}

::: {.fragment .fade-up}
**Mediation**

- Asks *how/why* X affects Y — through a mediator M on the causal path
- Three models: total effect (*c'*), X→M (*a* path), X+M→Y (*b* and *c* paths)
- Indirect effect = *a × b*; use `mediation::mediate()` for bootstrapped CIs
- Full mediation = direct effect disappears; partial = it shrinks
- Causal claims require strong theoretical justification with cross-sectional data
:::

::: {.fragment .fade-up}
**Multicollinearity**

- Occurs when predictors are highly correlated → inflated SEs, unstable coefficients
- Diagnose with `car::vif()`: VIF > 5 warrants attention, > 10 is severe
- Fix: drop redundant predictors, keep the theoretically central one
- Always report VIF in your paper (one sentence is enough)
:::

::: {.fragment .fade-up}
**Final paper**

- Bivariate + multivariate + one extension (interaction, mediation, or logistic)
- Use `modelsummary()` with `vcov = "HC1"` (robust SEs) throughout
- Lead with substance: what does the finding *mean* sociologically?
- A specific limitations paragraph signals methodological maturity
:::


## Why This Matters {.smaller}

::: {.fragment .fade-up}
Mediation is how sociology asks *mechanism* questions:

- Does education reduce poverty *because* it raises job quality? → mediation
- Does neighborhood segregation affect health *through* differential access to healthcare? → mediation
- Does discrimination reduce earnings *through* occupational sorting? → mediation
:::

::: {.fragment .fade-up}
These aren't just statistical questions — they're questions about **how social structures reproduce themselves**. Knowing the mechanism tells us *where to intervene*.
:::

::: {.fragment .fade-up}
::: {.callout-tip icon=false}
## The big picture
This course has moved from "is there a relationship?" (correlations, chi-squared) → "how strong is it?" (OLS) → "is it real?" (controls, confounding) → "for whom?" (interactions) → **"how does it work?"** (mediation). That's the arc of causal thinking in social science.
:::
:::


## Questions?


## Assignments {.smaller}

- HW 11 was canceled - wanted to make sure you had time to work on final exam 
- Will still be able to drop lowest of 10 assignments.

**Final Paper** — <span class="highlight">Due Thursday, May 7</span>

::: {.callout-note icon=false}
## Final paper components
1. Introduction and research question
2. Data and methods (describe attain.csv or your dataset, key variables, analytic strategy)
3. Results: bivariate model → multivariate model → extension (interaction or mediation)
4. Discussion: what do the results mean? connect to literature
5. Limitations: cross-sectional data, measurement issues, omitted variables
6. Model table using `modelsummary()` with robust SEs
7. At least one visualization (scatter + regression line, coefficient plot, or marginal effects plot)
:::

**In-class presentations** — <span class="highlight">Tuesday, April 28/May 5.</span>

- 7–10 minutes: question → data → results → so what
- Don't show the full model table to the class — use a coefficient plot or highlight 2–3 key numbers
