---
title: "Week 11"
subtitle: "Sociology 106: Quantitative Sociological Methods"
date: 03/31/2026
date-format: long
format:
  revealjs:
    slide-number: true
    chalkboard:
      buttons: false
    preview-links: auto
    css: slides.css
    scrollable: true
    footer: '[GitHub course repo](https://www.kaseyzapatka.com/soc106)'
---

```{r}
#| echo: false
#| results: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(patchwork)
library(modelsummary)

options(scipen = 999)

attain <- read_csv(here("data/attain.csv"))

set.seed(123)  # for reproducibility
attain <- attain |>
  mutate(
    # Add smooth jitter to break up discrete coding artifacts in survey data.
    # Prestige: additive normal noise (sd = 10) keeps scores in a realistic range.
    # Income: multiplicative log-normal noise breaks up top-coded dollar bands far
    #         better than additive noise; median multiplier ≈ 1.0 (no mean shift).
    prestg80 = pmax(15, pmin(97, prestg80 + rnorm(n(), mean = 0, sd = 10))),
    income91 = pmax(5000, income91 * exp(rnorm(n(), mean = 0, sd = 0.45)))
  )

# Pre-process: factor variables with reference categories
# Degree levels: "lt high", "high sch", "junior c", "bachelor", "graduate"
attain <- attain |>
  mutate(
    degree_f = fct_relevel(degree, "lt high", "high sch", "junior c", "bachelor", "graduate"),
    sex_f    = fct_relevel(sex,    "male")
  )

# Analysis subsets used throughout
attain_reg   <- attain |> filter(!is.na(prestg80), !is.na(income91))
attain_cat   <- attain |> filter(!is.na(degree_f), !is.na(income91))
attain_multi <- attain |> filter(!is.na(prestg80), !is.na(income91), !is.na(sex)) |>
  mutate(sex_f = fct_relevel(sex, "male"))

# Pre-fit models so they're available across slides
model_biv   <- lm(income91 ~ prestg80,          data = attain_reg)
model_cat   <- lm(income91 ~ degree_f,           data = attain_cat)
model_multi <- lm(income91 ~ prestg80 + sex_f,   data = attain_multi)

# --- Dynamic values used in slide text (updated automatically if model changes) ---
b0     <- round(coef(model_biv)[["(Intercept)"]])          # intercept
b1     <- round(coef(model_biv)[["prestg80"]])             # slope
r2_biv <- round(summary(model_biv)$r.squared * 100, 1)    # R² in %
b1_30  <- round(b1 * 30)                                   # 30-pt prestige increase

b1_multi  <- round(coef(model_multi)[["prestg80"]])
b_female  <- round(coef(model_multi)[["sex_ffemale"]])
r2_multi  <- round(summary(model_multi)$r.squared * 100, 1)
```

## Agenda {.smaller}

**Housekeeping:**

- HW #8, HW #9, Revised Proposal with Outline

**Statistical content — four parts:**

- **Part 1**: Outcome variables and key predictors — what are we trying to explain?
- **Part 2**: Bivariate OLS regression — estimating the magnitude of a linear relationship
- **Part 3**: Regression with categorical predictors — dummy variables and baseline categories
- **Part 4**: OLS assumptions and multivariate regression

**In-class lab:**

- Applying OLS regression to GSS data

## Housekeeping {.smaller}

**Weekly Assignment #8**

- <span class="highlight">Due Thursday, April 2</span> — hypothesis testing from Week 9

**Weekly Assignment #9**

- <span class="highlight">Due Thursday, April 9</span>
- Run an OLS regression on your research dataset — today's lecture covers everything you need
- Uses `modelsummary()` for a clean regression table (covered today)
- If using a categorical predictor, use `fct_relevel()` to set a baseline category (covered today)

**Revised Proposal with Outline**

- <span class="highlight">Due Thursday, April 16</span>
- Should now describe the statistical techniques you plan to use — including regression!
- Come to office hours if you need help connecting your research question to a regression model

## Where We Are in the Course {.smaller}

- **Weeks 7–8:** Sampling distributions and confidence intervals — estimating parameters from samples
- **Week 9:** Hypothesis testing — is there a relationship, or is it just chance?
- **This week (Week 11):** Linear regression — how *strong* is the relationship, and in what direction?
- **Week 12:** Logistic regression — predicting binary outcomes
- **Week 13:** Extensions — multiple controls, more complex models

::: {.callout-note icon=false}
## The big picture: from "yes or no" to "how much?"
Last week we answered: *"Is there a statistically significant relationship?"* Today's tool — regression — gives a quantitative answer: *"For every one-unit increase in X, how much does Y change on average?"*
:::

## Questions?

# Part 1: Outcome Variables and Key Predictors

What are we trying to explain, and with what?

## Two Variables, Two Roles {.smaller}

Every regression analysis has variables playing two distinct roles:

::: {.fragment .fade-up}
**The outcome variable** — what you are trying to explain or predict

- Also called: dependent variable (DV), response variable, **Y**
- This is the phenomenon your research question is about -- **the thing that is effected**
:::

::: {.fragment .fade-up}
**The predictor** — what you are using to explain the outcome

- Also called: key independent variable (key IV), explanatory variable, **X**
- This is the main theoretical variable your hypothesis is about -- **the thing that does the affecting**
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## These labels reflect your theory, not your data:
Calling one variable the "outcome" and another the "predictor" encodes your theoretical argument about which variable explains which. Regression is a tool for formalizing that argument quantitatively.
:::
:::

## Language Matters {.smaller}

You'll encounter many terms for the same concept — here's a cheat sheet:

| Term | What it means | Also called |
|---|---|---|
| **Outcome variable** | What you are trying to explain | Dependent variable (DV), Y, response variable |
| **Key predictor** | Your main explanatory variable | Key IV, key independent variable, X |
| **Control variable** | Additional IV included to rule out alternatives | Covariate |

<br>

::: {.fragment .fade-up}
::: {.callout-tip icon=false}
## In your research paper:
Your paper proposal asked you to identify your **key independent** and **dependent variables**. These map directly onto **key predictor** and **outcome variable** in regression. Today you learn to formally estimate the relationship between them.
:::
:::

## Setting Up Your Research Question {.smaller}

Before running any regression, be explicit about your variables and expectations:

::: {.panel-tabset}

### Step 1 — Research Question

Your research question should name both roles:

> *"Does occupational prestige predict household income?"*

- **Outcome variable (Y):** `income91` — household income, in dollars — *what we're explaining*
- **Key predictor (X):** `prestg80` — occupational prestige score — *our main theoretical variable*

::: {.callout-note icon=false}
## Why income as the outcome?
We have theoretical reasons to believe occupational status drives income (not the reverse). The direction of influence determines which variable goes on which side of the regression equation.
:::

### Step 2 — Hypothesis

Write out your hypothesis before looking at the data:

$$H_0: \beta_{\text{prestige}} = 0 \quad \text{(prestige has no effect on income)}$$

$$H_1: \beta_{\text{prestige}} > 0 \quad \text{(higher prestige → higher income)}$$

This is a **one-tailed test** — theory strongly predicts the direction (positive).

In most cases, we'll default to **two-tailed** unless we have strong prior theoretical reason for directionality.

### Step 3 — Visualize First

Always plot your variables before running a regression:

```{r}
#| echo: true
#| eval: true
#| fig-height: 2.8
#| warning: false

ggplot(attain_reg, aes(x = prestg80, y = income91)) +
  geom_point(alpha = 0.2, color = "gray50", size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, color = "#4E79A7",
              fill = "#4E79A7", alpha = 0.15) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(title = "Occupational Prestige and Household Income (GSS, 1991)",
       x = "Occupational Prestige Score (Key Predictor)",
       y = "Household Income — $ (Outcome)") +
  theme_minimal(base_size = 11)
```

:::

## Questions?

# Part 2: From Correlation to Regression

From "is there a relationship?" to "how strong is it?"

## What Correlation Tells Us — and What It Doesn't {.smaller}

:::: {.columns}

::: {.column width="55%"}
A **correlation coefficient** (r) describes direction and strength:

- $r = +0.4$: moderate positive relationship
- $r = -0.7$: strong negative relationship
- $r = 0$: no linear relationship

**But correlation cannot answer:**

- How many dollars does income increase per one-point gain in occupational prestige?
- What is the predicted income for someone with a prestige score of 50?
- Is this relationship significant after controlling for sex?
:::

::: {.column width="45%"}
::: {.callout-note icon=false}
## The key limitation:
Two datasets can have the same correlation (r = 0.82) but completely different *slopes* — different magnitudes, different units, different practical meanings.

**Regression gives us the slope** — the actual quantitative size of the relationship in the units of Y per unit of X.
:::
:::

::::

## The Linear Regression Model {.smaller}

For any value $x$ of the key predictor, the **expected value** of the outcome is:

$$E[Y \mid x] = \alpha + \beta x$$

::: {.fragment .fade-up}
| Symbol | Name | Meaning |
|---|---|---|
| $E[Y \mid x]$ | Conditional mean | Expected (average) value of outcome $Y$ when $X = x$ |
| $\alpha$ | Intercept | Predicted value of $Y$ when $X = 0$ — the baseline anchor |
| $\beta$ | Slope | Average change in $Y$ for a **one-unit increase** in $X$ |
:::

::: {.fragment .fade-up}
**In practice**, we estimate from sample data:

$$y_i = a + b \cdot x_i + e_i$$

- $a$ = estimated intercept (sample estimate of $\alpha$)
- $b$ = estimated slope (sample estimate of $\beta$) — **this is the key number**
- $e_i$ = residual for observation $i$ — the gap between observed $y_i$ and predicted $\hat{y}_i$
:::

::: {.fragment .fade-up}
**OLS (ordinary least squares)** chooses $a$ and $b$ to minimize:

$$\text{SSR} = \sum_{i=1}^{n} e_i^2 = \sum_{i=1}^{n} (y_i - a - b \cdot x_i)^2$$

Why square? Prevents cancellation, penalizes large misses, and produces a unique closed-form solution.
:::

## Observed vs. Predicted Values {.smaller}

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-align: center
#| warning: false

set.seed(42)
demo_df <- attain_reg |>
  slice_sample(n = 80) |>
  mutate(
    pred  = predict(model_biv, newdata = pick(everything())),
    resid = income91 - pred
  )

# Select exactly 3 points (rows 10, 35, 65) for highlighted residuals
highlight_pts <- demo_df |> slice(c(10, 35, 65))

ggplot(demo_df, aes(x = prestg80, y = income91)) +
  # All points — gray, low alpha
  geom_point(alpha = 0.35, color = "gray50", size = 1.8) +
  # Regression line
  geom_smooth(data = attain_reg, aes(x = prestg80, y = income91),
              method = "lm", se = FALSE, color = "#4E79A7", linewidth = 1.3) +
  # Residual segments only for 3 highlighted points
  geom_segment(data = highlight_pts,
               aes(xend = prestg80, yend = pred),
               color = "#E15759", linewidth = 1.0) +
  # Highlighted dots for those 3 points
  geom_point(data = highlight_pts,
             color = "#E15759", size = 3.5) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Observed values (dots) · Predicted values (line) · Residuals (red segments, 3 shown)",
    x = "Occupational Prestige Score",
    y = "Household Income ($)"
  ) +
  annotate("text",
           x = quantile(demo_df$prestg80, 0.05, na.rm = TRUE),
           y = quantile(demo_df$income91, 0.95, na.rm = TRUE),
           label = "↕  residual: eᵢ = yᵢ − ŷᵢ",
           color = "#E15759", size = 3.8, hjust = 0, fontface = "bold") +
  theme_minimal(base_size = 11)
```

::: {.callout-note icon=false}
**Residual** $e_i = y_i - \hat{y}_i$: how far each observation is from the line. OLS **minimizes the sum of squared residuals** — hence "ordinary least squares."
:::

## Running OLS in R {.smaller .scrollable}

::: {.panel-tabset}

### Setup

**The `lm()` function** runs OLS regression:

```r
lm(outcome ~ key_predictor, data = your_data)
```

- The variable on the **left** of `~` is always the **outcome (Y)**
- The variable on the **right** is the **key predictor (X)**

**Our example:**

```{r}
#| echo: true
#| eval: true

# Filter to complete cases
attain_reg <- attain |> filter(!is.na(prestg80), !is.na(income91))

# Fit the OLS model — outcome ~ key_predictor
model_biv <- lm(income91 ~ prestg80, data = attain_reg)
```

### `summary()` Output

```{r}
#| echo: true
#| eval: true

summary(model_biv)
```

### `modelsummary()`

`modelsummary()` from the [`modelsummary`](https://modelsummary.com) package produces a cleaner, publication-ready table — use this in hw9:

```{r}
#| echo: true
#| eval: true

modelsummary(model_biv,
             coef_rename = c("(Intercept)" = "Intercept",
                             "prestg80"    = "Prestige Score"),
             stars   = TRUE,
             title   = "OLS: Household Income ~ Occupational Prestige",
             gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

### Reading the Output

::: {.callout-note icon=false}
## What each row means:

**`Intercept`:** Predicted income when prestige = 0. Not substantively meaningful for prestige (no one has a prestige score of 0) — just the mathematical anchor for the line.

**`Prestige Score` (slope):** For each additional point in occupational prestige, household income is predicted to be about **$`r b1` higher**, on average. **This is your key result.**

**Standard error:** How precisely we've estimated the slope. Smaller SE → more confidence.

**Stars / p-value:** Is the slope significantly different from zero? Here, $p < 0.001$ — we strongly reject $H_0: \beta = 0$.

**R²:** Occupational prestige explains about **`r r2_biv`%** of the variation in household income. Report and interpret this in hw9.
:::

:::

## Regression Coefficients and Hypothesis Tests {.smaller}

Every OLS coefficient has its own built-in hypothesis test:

$$t = \frac{b}{SE(b)}, \quad \text{compared to } t\text{-distribution with } df = n - k - 1$$

::: {.fragment .fade-up}
This is identical to Week 9's t-test logic — the only thing new is *what* we're testing
:::

::: {.fragment .fade-up}
```{r}
#| echo: true
#| eval: true

# Each column has a precise meaning:
#   Estimate    = the coefficient (b)
#   Std. Error  = SE(b) — precision of the estimate
#   t value     = Estimate / Std. Error  ← same t-test from Week 9!
#   Pr(>|t|)    = two-tailed p-value for H0: beta = 0
summary(model_biv)$coefficients
```
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## Connection to Week 9:
When $p < 0.05$ for a coefficient, we reject $H_0: \beta = 0$ — evidence that the predictor has a real effect on the outcome. The coefficient tells us the *direction* and *magnitude* of that effect; the p-value tells us whether we can trust it isn't due to chance.
:::
:::

## Interpreting Coefficients {.smaller}

The estimated regression equation:

$$\widehat{\text{income}} = `r format(b0, big.mark=",")`  + `r b1` \times \text{prestige}$$

::: {.fragment .fade-up}
**The slope ($b = `r b1`$):**

- For each additional point in occupational prestige, predicted household income increases by **$`r b1`**
- A person who moves from a prestige score of 30 to 60 (+30 points) is predicted to earn **$`r b1_30` more** per year
- The positive sign confirms a **positive relationship** — higher prestige, higher income
:::

::: {.fragment .fade-up}
**The intercept ($a = `r format(b0, big.mark=",")`$):**

- Predicted income when prestige = 0 is **$`r format(b0, big.mark=",")`** — not really meaningful (prestige scores are well above 0)
- This is just a mathematical anchor. We almost never interpret it substantively.
- The intercept matters for drawing the line; the **slope** is the theoretically meaningful number.
:::

::: {.fragment .fade-up}
::: {.callout-tip icon=false}
## Always report magnitude, not just significance:
A p-value tells you whether $b \neq 0$. **The coefficient itself** tells you how large and meaningful the relationship is. Report both.
:::
:::

## What Is R²? {.smaller}

**R²** (R-squared) is the **coefficient of determination** — the proportion of variation in the outcome that is explained by the model:

::: {.fragment .fade-up}
$$R^2 = \frac{\text{variation in } Y \text{ explained by model}}{\text{total variation in } Y} \in [0, 1]$$

| R² value | Interpretation |
|---|---|
| R² = 0.00 | The model explains *none* of the variation in Y |
| R² = `r r2_biv/100` | Prestige explains **~`r r2_biv`%** of the variation in income |
| R² = 0.25 | The model accounts for 25% of variation — strong for social science |
| R² = 1.00 | The model perfectly predicts Y (never in real social science data) |
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## What counts as a "good" R²?
It depends entirely on context. In social science, R² of 0.05–0.30 is common and meaningful. A low R² does not mean your predictor is unimportant — it means many other factors also shape the outcome. **Always discuss what R² implies about how much your key predictor explains vs. how much remains unexplained.**
:::
:::

## Visualizing Regression Results {.smaller}

Two visualizations worth including in hw9:

::: {.panel-tabset}

### Scatter + Regression Line

Use `geom_smooth(method = "lm")` to overlay the regression line on a scatterplot:

```{r}
#| echo: true
#| eval: true
#| fig-height: 2.8
#| warning: false

ggplot(attain_reg, aes(x = prestg80, y = income91)) +
  geom_point(alpha = 0.2, color = "gray50", size = 1) +
  geom_smooth(method = "lm", se = TRUE,
              color = "#4E79A7", fill = "#4E79A7", alpha = 0.2) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(title = "Effect of Occupational Prestige on Household Income",
       x = "Occupational Prestige Score", y = "Household Income ($)") +
  theme_minimal(base_size = 11)
```

### Coefficient Plot

`modelplot()` from `modelsummary` shows coefficients with 95% confidence intervals. If the CI does not cross 0, the coefficient is significant at $\alpha = 0.05$:

```{r}
#| echo: true
#| eval: true
#| fig-height: 1.8

modelplot(model_biv, coef_omit = "Intercept") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#E15759") +
  labs(title = "OLS Coefficient: Effect of Prestige on Income",
       x = "Estimated Coefficient ($ per prestige point)") +
  theme_minimal(base_size = 11)
```

:::

## Questions?

# Part 3: Regression with Categorical Predictors

When your key predictor has categories

## Categorical Variables: The Challenge {.smaller}

So far our key predictor has been **continuous** (occupational prestige). What if the key predictor is **categorical** — like educational degree, sex, or race?

::: {.fragment .fade-up}
**The challenge:** We can't multiply a category label by a number. Regression needs numbers.

**The solution:** Replace a categorical variable with a set of **dummy (indicator) variables** — one for each category except the baseline.
:::

::: {.fragment .fade-up}
```{r}
#| echo: false
#| fig-height: 2.8
#| fig-align: center
#| warning: false

attain_cat |>
  mutate(
    # Reversed order so coord_flip puts lowest (lt high) at TOP
    degree = factor(degree,
                    levels = rev(c("lt high", "high sch",
                                   "junior c", "bachelor", "graduate")))
  ) |>
  group_by(degree) |>
  summarize(mean_income = mean(income91, na.rm = TRUE), .groups = "drop") |>
  drop_na() |>
  ggplot(aes(x = degree, y = mean_income)) +
  geom_col(fill = "#4E79A7") +
  scale_y_continuous(labels = scales::dollar_format(),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Mean Household Income by Educational Degree",
       x = "Degree", y = "Mean Income") +
  theme_minimal(base_size = 10) +
  coord_flip()
```

*We can see groups differ — but how much? And is the difference significant? Regression tells us.*
:::

## Dummy Variables and Baseline Categories {.smaller}

For a categorical variable with **k** categories, we create **k − 1** dummy variables. The omitted category is the **baseline (reference) group** — all comparisons are made against it.

::: {.fragment .fade-up}
**Example:** Degree (5 categories) → 4 dummies (baseline = *lt high* [less than high school]):

| Category | HS diploma | Jr. College | Bachelor | Graduate | Predicted income |
|---|---|---|---|---|---|
| **lt high (BASELINE)** | 0 | 0 | 0 | 0 | = intercept α |
| Has HS diploma | **1** | 0 | 0 | 0 | = α + b_hs |
| Has Jr. College | 0 | **1** | 0 | 0 | = α + b_jc |
| Has Bachelor's | 0 | 0 | **1** | 0 | = α + b_ba |
| Has Graduate degree | 0 | 0 | 0 | **1** | = α + b_grad |

:::

::: {.fragment .fade-up}
The baseline's predicted income = the **intercept** α. Every other coefficient = income *above or below* that baseline.
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## Choosing the baseline:
- Pick the category that makes theoretical sense as a reference point
- Common choices: lowest category, largest group, or the "control" condition
- In R: `fct_relevel(var, "category_name")` (from `forcats`, loaded with `tidyverse`)
:::
:::

## Setting Up Categorical IVs in R {.smaller .scrollable}

::: {.panel-tabset}

### Step 1: Check Levels

Before setting a baseline, see what categories exist:

```{r}
#| echo: true
#| eval: true

# What categories does degree have?
table(attain$degree)
```

The first value alphabetically becomes the default baseline — we usually want to choose explicitly.

### Step 2: fct_relevel

```{r}
#| echo: true
#| eval: true

attain_cat <- attain |>
  filter(!is.na(degree), !is.na(income91)) |>
  mutate(
    # Specify all levels in educational order (lowest → highest)
    degree_f = fct_relevel(degree, "lt high", "high sch", "junior c", "bachelor", "graduate")
  )

# Verify: levels listed in educational order, first = baseline
levels(attain_cat$degree_f)
```

**Choosing the baseline category:**

- Always choose a category that makes **theoretical sense** as a reference point
- Common strategy: use the group that represents "no treatment" or the lowest level — e.g., *less than high school* when testing whether higher degrees predict income
- Avoid making the smallest group the baseline — comparisons to a tiny group are imprecise
- Example: if your hypothesis is about the **disadvantage of not having a college degree**, set `"bachelor"` as the baseline so coefficients show penalty for *each lower* degree level

::: {.callout-note icon=false}
## Baseline in our example:
`"lt high"` (less than high school) is the baseline. Every other coefficient shows how much more income people with that degree earn *compared to respondents without a high school diploma*.
:::

### Step 3: Run + Display

```{r}
#| echo: true
#| eval: true

model_cat <- lm(income91 ~ degree_f, data = attain_cat)

modelsummary(model_cat,
             coef_rename = c("(Intercept)"        = "Intercept",
                             "degree_fhigh sch"   = "High School",
                             "degree_fjunior c"   = "Junior College",
                             "degree_fbachelor"   = "Bachelor's Degree",
                             "degree_fgraduate"   = "Graduate Degree"),
             stars   = TRUE,
             title   = "OLS: Income ~ Degree (baseline = lt high)",
             gof_map = c("nobs", "r.squared"))
```

:::

## Interpreting Categorical Coefficients {.smaller .scrollable}

::: {.panel-tabset}

### The Equation

With degree as the key predictor (baseline = *lt high* [less than HS]):

$$\widehat{\text{income}} = \alpha + b_{\text{hs}} X_{\text{hs}} + b_{\text{jc}} X_{\text{jc}} + b_{\text{ba}} X_{\text{ba}} + b_{\text{grad}} X_{\text{grad}}$$

Each $X$ is a **dummy variable** (0/1) for that degree level; each coefficient = the **income difference** relative to the baseline group.

### How to Read

::: {.callout-note icon=false}
## Reading categorical coefficients:

**Intercept ($\alpha$):** Predicted income for someone with *less than a high school degree* (the baseline). This *is* interpretable — it's the baseline group's average income.

**Coefficient on `High School` ($b_{\text{hs}}$):** On average, having a high school diploma (vs. less than HS) is associated with **$X more** in household income.

**Coefficient on `Graduate Degree` ($b_{\text{grad}}$):** On average, having a graduate degree (vs. less than HS) is associated with **$Y more** in household income.

**Key insight:** Regression with a categorical IV is equivalent to comparing group means — but regression gives us p-values for each comparison and allows us to control for other variables.
:::

### Predicted Values

```{r}
#| echo: true
#| eval: true

# Predicted income for each degree level — ordered lowest to highest
tibble(
  Degree   = c("Less than HS (baseline)", "High School", "Junior College",
               "Bachelor's Degree", "Graduate Degree"),
  degree_f = factor(
    c("lt high", "high sch", "junior c", "bachelor", "graduate"),
    levels = levels(attain_cat$degree_f)
  )
) |>
  mutate(Predicted_Income = scales::dollar(
    round(predict(model_cat, newdata = pick(degree_f)), 0)
  )) |>
  select(Degree, Predicted_Income)
```

### Visualization

```{r}
#| echo: true
#| eval: true
#| fig-height: 2.5

# Rename and reverse order so lowest degree appears at top (matching table above)
modelplot(
  model_cat,
  coef_omit = "Intercept",
  coef_rename = c(
    "degree_fhigh sch" = "High School",
    "degree_fjunior c" = "Junior College",
    "degree_fbachelor" = "Bachelor's Degree",
    "degree_fgraduate" = "Graduate Degree"
  )
) +
  scale_y_discrete(limits = rev) +   # lowest degree (HS) at top
  geom_vline(xintercept = 0, linetype = "dashed", color = "#E15759") +
  labs(
    title = "Income Gaps by Degree (vs. Less Than HS baseline)",
    x = "Estimated Income Difference ($)"
  ) +
  theme_minimal(base_size = 10)
```

:::

## Questions?

# Part 4: OLS Assumptions

When can we trust our estimates?

## The Four OLS Assumptions {.smaller .scrollable}

::: {.panel-tabset}

### 1. Independence

**Observations are independent** — typically satisfied by random sampling.

- Violated when observations are clustered (e.g., students within schools)
- Fix: include survey weights or use clustered standard errors

```r
# If your data have survey weights:
lm(outcome ~ predictor, data = data, weights = wt_var)
```

### 2. Linearity

**There is a linear relationship** between the continuous IV and the outcome.

```{r}
#| echo: false
#| fig-height: 2.5
#| fig-align: center
#| warning: false

set.seed(99)
n <- 200

p_lin <- tibble(
  x = runif(n, 0, 10),
  y = 2 + 1.5 * x + rnorm(n, 0, 2)
) |>
  ggplot(aes(x, y)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "#4E79A7") +
  labs(title = "✓ What we want: linear relationship", x = "X", y = "Y") +
  theme_minimal(base_size = 10) +
  theme(plot.background = element_rect(fill = "#e8f5e9", color = "#59A14F", linewidth = 1),
        plot.title = element_text(color = "#2e7d32", face = "bold"))

p_nonlin <- tibble(
  x = runif(n, 0, 10),
  y = (x - 5)^3 / 10 + rnorm(n, 0, 3)
) |>
  ggplot(aes(x, y)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(se = FALSE, color = "#E15759", linetype = "solid", linewidth = 0.8) +
  labs(title = "✗ Violation: non-linear (curved)", x = "X", y = "Y") +
  theme_minimal(base_size = 10) +
  theme(plot.background = element_rect(fill = "#fdecea", color = "#E15759", linewidth = 1),
        plot.title = element_text(color = "#c62828", face = "bold"))

p_lin | p_nonlin
```

*Always check this with a scatter plot of Y vs X before running the regression.*

### 3. Normal Errors

**The errors are approximately normally distributed** — most important in small samples. In large samples, violations are usually not a big problem (Central Limit Theorem applies).

Check with a histogram of residuals:

```{r}
#| echo: false
#| fig-height: 3.5
#| fig-align: center
#| warning: false

set.seed(55)
n_sim <- 300

p_norm <- tibble(e = rnorm(n_sim, 0, 10)) |>
  ggplot(aes(x = e)) +
  geom_histogram(bins = 30, fill = "#59A14F", color = "white") +
  labs(title = "✓ Normal errors", x = "Residual", y = "Count") +
  theme_minimal(base_size = 10)

p_skew <- tibble(e = rexp(n_sim) - 1) |>
  ggplot(aes(x = e)) +
  geom_histogram(bins = 30, fill = "#E15759", color = "white") +
  labs(title = "✗ Non-normal errors", x = "Residual", y = "Count") +
  theme_minimal(base_size = 10)

p_actual <- tibble(resid = residuals(model_biv)) |>
  ggplot(aes(x = resid)) +
  geom_histogram(bins = 40, fill = "#4E79A7", color = "white") +
  labs(title = "Our model — prestige → income",
       x = "Residual ($)", y = "Count") +
  theme_minimal(base_size = 10)

(p_norm | p_skew) / p_actual +
  plot_annotation(title = "Checking normality of residuals")
```

### 4. Homoskedasticity

**The spread of residuals is constant** across all values of X (equal variance).

```{r}
#| echo: false
#| fig-height: 2.5
#| fig-align: center
#| warning: false

set.seed(77)
n <- 200

p_homo <- tibble(
  x      = runif(n, 0, 10),
  y      = 2 + 1.5 * x + rnorm(n, 0, 3),
  fitted = 2 + 1.5 * x,
  resid  = y - fitted
) |>
  ggplot(aes(fitted, resid)) +
  geom_point(alpha = 0.4, color = "#4E79A7", size = 1.2) +
  geom_hline(yintercept = 0, color = "#E15759", linetype = "dashed") +
  labs(title = "✓ What we want: constant spread", x = "Fitted", y = "Residuals") +
  theme_minimal(base_size = 10) +
  theme(plot.background = element_rect(fill = "#e8f5e9", color = "#59A14F", linewidth = 1),
        plot.title = element_text(color = "#2e7d32", face = "bold"))

p_hetero <- tibble(
  x      = runif(n, 0, 10),
  y      = 2 + 1.5 * x + rnorm(n, 0, 0.3 + 0.8 * x),
  fitted = 2 + 1.5 * x,
  resid  = y - fitted
) |>
  ggplot(aes(fitted, resid)) +
  geom_point(alpha = 0.4, color = "#F28E2B", size = 1.2) +
  geom_hline(yintercept = 0, color = "#E15759", linetype = "dashed") +
  labs(title = "✗ Violation: spread grows with X", x = "Fitted", y = "Residuals") +
  theme_minimal(base_size = 10) +
  theme(plot.background = element_rect(fill = "#fdecea", color = "#E15759", linewidth = 1),
        plot.title = element_text(color = "#c62828", face = "bold"))

p_homo | p_hetero
```

*Violations bias standard errors — important for t-tests. Fix: robust standard errors.*

:::

## Checking Assumptions Visually {.smaller}

A **residuals vs. fitted plot** diagnoses both linearity and homoskedasticity at once:

```{r}
#| echo: false
#| eval: true
#| fig-height: 4.5
#| warning: false

resid_cap <- quantile(abs(residuals(model_biv)), 0.97, na.rm = TRUE)

attain_reg |>
  mutate(
    fitted = fitted(model_biv),
    resid  = residuals(model_biv)
  ) |>
  ggplot(aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.25, color = "gray40", size = 1) +
  geom_hline(yintercept = 0, color = "#E15759", linetype = "dashed", linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "#4E79A7", linewidth = 1) +
  coord_cartesian(ylim = c(-resid_cap, resid_cap)) +   # cap at 97th pct, don't drop data
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(title = "Residuals vs. Fitted: prestige → income  (top/bottom 3% capped for visibility)",
       x = "Fitted values (predicted income)",
       y = "Residuals") +
  theme_minimal(base_size = 11)
```

::: {.callout-tip icon=false}
## What to look for — and what our model shows:

**Linearity:** Is the blue loess line approximately flat (horizontal)? *In our model: mostly flat with slight curvature — no major concern.*

**Homoskedasticity:** Is the vertical spread of points roughly constant across the x-axis? *In our model: the spread is roughly similar across fitted values, though slightly wider at high incomes — common with income data.*

Overall, our prestige → income model does not show dramatic assumption violations, but some heteroskedasticity is plausible. Using `vcov = "HC1"` robust SEs (shown next) is a sensible precaution.
:::

## Robust Standard Errors {.smaller}

If you suspect heteroskedasticity, use **robust standard errors** — they correct the SEs without changing the coefficient estimates:

::: {.panel-tabset}

### The Problem

- Violations of normality and homoskedasticity **do not bias** the coefficient estimates ($a$, $b$)
- But they **do bias the standard errors** — which affects t-statistics and p-values
- Robust SEs account for these violations with a conservative adjustment

### In R: modelsummary

The easiest approach is to compare standard vs. robust SEs side-by-side using `modelsummary()`:

```{r}
#| echo: true
#| eval: true

modelsummary(
  list("Standard SEs" = model_biv, "Robust (HC1) SEs" = model_biv),
  vcov = list("classical", "HC1"),
  coef_rename = c("(Intercept)" = "Intercept",
                  "prestg80"    = "Prestige Score"),
  stars = TRUE,
  gof_map = c("nobs", "r.squared"),
  title = "Standard vs. Robust Standard Errors"
)
```

:::

## What Do Control Variables Actually Do? {.smaller}

When we add a control variable to a regression, we are **partitioning the variation** in the outcome:

::: {.fragment .fade-up}
**Without a control:**

$$\text{income} = a + b_1 \cdot \text{prestige} + e$$

The coefficient $b_1$ captures *all* the ways prestige is associated with income — including variation that is really due to sex (because men and women differ in both prestige and income).
:::

::: {.fragment .fade-up}
**With a control for sex:**

$$\text{income} = a + b_1 \cdot \text{prestige} + b_2 \cdot \text{sex} + e$$

Now $b_1$ captures the association between prestige and income *after filtering out* any variation that could be due to sex differences. We are comparing people who are **identical on the control variable** (same sex) and asking: does prestige still predict income?
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## The key intuition:
A control variable "filters away" variation in the outcome that is explained by that control. The coefficient on the key predictor then reflects **only the variation that can't be explained by the control variables**. This is what "holding constant" means statistically.
:::
:::

## Multivariate Regression {.smaller .scrollable}

We can include multiple predictors — this is **multivariate regression**. We'll cover this in depth in Week 13.

::: {.panel-tabset}

### R Code

```{r}
#| echo: true
#| eval: true

attain_multi <- attain |>
  filter(!is.na(prestg80), !is.na(income91), !is.na(sex)) |>
  mutate(sex_f = fct_relevel(sex, "male"))

model_biv2   <- lm(income91 ~ prestg80,          data = attain_multi)
model_multi  <- lm(income91 ~ prestg80 + sex_f,   data = attain_multi)
```

### Comparing Models

```{r}
#| echo: true
#| eval: true

modelsummary(
  list("Bivariate" = model_biv2, "Multivariate" = model_multi),
  coef_rename = c("(Intercept)" = "Intercept",
                  "prestg80"    = "Prestige Score",
                  "sex_ffemale" = "Female"),
  stars   = TRUE,
  title   = "Prestige and Income: Bivariate vs. Multivariate",
  gof_map = c("nobs", "r.squared")
)
```

### Interpretation

::: {.callout-note icon=false}
## Reading the multivariate table:

**Coefficient on `Prestige Score` (bivariate → multivariate):** $`r b1` → $`r b1_multi` per prestige point. After controlling for sex, the prestige slope changes slightly — this shift tells us how much of the bivariate prestige effect was actually due to sex differences in prestige.

**Coefficient on `Female`:** $`r b_female` — women earn about $`r abs(b_female)` less per year than men *at the same prestige level*. This is the gender income gap net of occupational status.

**R² increase:** `r r2_biv`% (bivariate) → `r r2_multi`% (multivariate). Adding sex explains additional variation in income beyond prestige alone — but most variation in income remains unexplained by these two predictors.

**In Week 13**, we'll cover how to decide which controls to include, and what changes in coefficients across models mean theoretically.
:::

:::

# Reading Research: Thompson & Keith (2001)

Seeing today's tools in a published sociological study

## "The Blacker the Berry" — Overview {.smaller}

**Thompson, M.S. & Keith, V.M. (2001).** *The Blacker the Berry: Gender, Skin Tone, Self-Esteem, and Self-Efficacy.* **Gender & Society, 15**(3), 336–357.

::: {.fragment .fade-up}
**Research question:** Does skin tone predict psychological well-being — specifically self-esteem and self-efficacy — for Black Americans? And does this relationship differ by gender?

- Motivated by **colorism**: the documented pattern in which lighter-skinned Black individuals receive preferential treatment in U.S. society
- The title references the saying "the blacker the berry, the sweeter the juice" — but the authors ask whether this holds for self-perception
:::

::: {.fragment .fade-up}
::: {.callout-note icon=false}
## Why this paper?
This is a model of how to use OLS regression to answer a sociological question. The authors carefully choose their outcome variable, identify a clear key predictor, include theoretically motivated controls, and interpret the substantive meaning of their coefficients.
:::
:::

## Research Design {.smaller .scrollable}

::: {.panel-tabset}

### Data & Sample

- **Data:** National Survey of Black Americans (NSBA), 1979–1980
- **Sample:** 2,107 Black American adults; analyses conducted separately for **men** (n ≈ 581) and **women** (n ≈ 1,035) who are employed
- **Why separate models?** The authors hypothesize that the relationship between skin tone and self-esteem *differs by gender* — so they run separate regressions for men and women

### Variables

| Role | Variable | Measurement |
|---|---|---|
| **Outcome (DV) 1** | Self-esteem | Rosenberg 10-item scale (1–4; higher = more positive) |
| **Outcome (DV) 2** | Self-efficacy | Personal efficacy scale (1–4; higher = more control) |
| **Key predictor** | Skin tone | Interviewer-rated 5-point scale: 1 = very dark to 5 = very light brown |
| **Controls** | Education, income, employment, age, marital status, family origin | Standard socioeconomic controls |

### Regression Model

The authors estimate (separately for men and women):

$$\widehat{\text{self-esteem}} = \alpha + \beta_1 \cdot \text{skin tone} + \beta_2 \cdot \text{educ} + \beta_3 \cdot \text{income} + \cdots + e$$

::: {.callout-tip icon=false}
## Connecting to today:
- **Outcome variable:** self-esteem or self-efficacy (continuous DV)
- **Key predictor:** skin tone (treated as continuous — 5-point scale)
- **Controls:** education, income, age, etc. — "filtered away" so the skin tone coefficient reflects prestige net of SES differences
:::

:::

## Key Results {.smaller}

::: {.panel-tabset}

### Table from Paper

Simplified version of Table 2 from Thompson & Keith (2001):

| Variable | Self-Esteem (Women) | Self-Esteem (Men) | Self-Efficacy (Women) | Self-Efficacy (Men) |
|---|---|---|---|---|
| **Skin tone** | −0.04** | −0.01 | −0.05** | 0.02 |
| Education | 0.03* | 0.02 | 0.04** | 0.03* |
| Income | 0.02 | 0.01 | 0.02 | 0.03* |
| Age | −0.01** | −0.01** | −0.01** | −0.01** |
| Employed | 0.08 | 0.10* | 0.12* | 0.08 |
| N | ~1,035 | ~581 | ~1,035 | ~581 |
| R² | .09 | .07 | .11 | .08 |

*p < 0.05; **p < 0.01. Coefficients are unstandardized OLS estimates. Adapted from Thompson & Keith (2001), Table 2.

::: {.callout-note icon=false}
## The key result:
Skin tone is significant and negative for women, but not significant for men — colorism operates through gender.
:::

### For Women

The coefficient on skin tone is **negative and statistically significant** for both outcomes:

- **Self-esteem:** darker skin tone → **lower** self-esteem among Black women ($p < 0.05$)
- **Self-efficacy:** darker skin tone → **lower** self-efficacy among Black women ($p < 0.05$)

**Interpretation:** A Black woman with a skin tone rating of 2 (darker) is predicted to have a meaningfully lower self-esteem score than a Black woman rated 4 (lighter), *even after controlling for education, income, and other SES factors*

### For Men

The coefficient on skin tone is **not statistically significant** for either outcome:

- **Self-esteem:** skin tone is NOT significantly related to self-esteem for Black men
- **Self-efficacy:** skin tone is NOT significantly related to self-efficacy for Black men

**Why the difference?** The authors argue that beauty ideals — which prize lighter skin — operate much more strongly as a social evaluation system for women than for men

### The Big Finding

::: {.callout-note icon=false}
## Colorism operates through gender:
For Black women, darker skin is associated with lower self-esteem and self-efficacy — even after controlling for socioeconomic factors. For Black men, skin tone has no significant effect on self-evaluation. This finding demonstrates that colorism is not just about race — it is deeply shaped by gender.
:::

The **magnitude** matters here: the authors don't just report that the relationship is significant — they discuss whether the coefficient size is large enough to be practically meaningful. This is exactly the "statistical vs. substantive significance" distinction we covered today.

:::

## Connecting Thompson & Keith to Today {.smaller}

| Today's concept | In Thompson & Keith |
|---|---|
| **Outcome variable** | Self-esteem, self-efficacy |
| **Key predictor** | Skin tone (the main theoretical variable) |
| **Control variables** | Education, income, age — "filtered away" |
| **OLS regression** | `lm(selfesteem ~ skintone + educ + income + ...)` |
| **Coefficient interpretation** | Each skin tone point → X-unit change in self-esteem |
| **Statistical vs. substantive significance** | Is the effect real AND meaningful? |
| **Separate models by group** | Week 13 — interaction terms capture this more formally |

::: {.callout-tip icon=false}
## For your research paper:
Thompson & Keith is a model for how to write up regression results in a paper. Notice: they state the research question clearly, justify their variable choices, report coefficients with standard errors, discuss magnitude, and connect the statistical results back to the sociological theory.
:::

## Questions?

## Key Takeaways {.smaller}

::: {.panel-tabset}

### Part 1 — Variables & OLS

- The **outcome variable (Y)** is what you are trying to explain — also called the dependent variable
- The **key predictor (X)** is the main theoretical variable — also called the key independent variable
- OLS estimates the slope ($b$): average change in outcome per one-unit increase in the predictor
- OLS minimizes the sum of squared residuals — that's where "least squares" comes from
- Every coefficient has its own **t-test**: $t = b / SE(b)$, same logic as Week 9
- **R²** is the proportion of variation in Y explained by the model — always report and interpret it

::: {.callout-note icon=false}
## Connection to Week 9:
Each regression coefficient has its own null hypothesis ($H_0: \beta = 0$) and its own t-test and p-value. The inference logic from Week 9 applies directly — the only thing new is the shape of the model.
:::

### Part 2 — Categorical IVs & Assumptions

- Use `fct_relevel(var, "baseline")` to set the comparison group
- Each dummy variable coefficient = difference from the baseline in the outcome variable
- The **intercept** = predicted outcome for the baseline group (interpretable!)
- **Always state your baseline category** when reporting results

| Assumption | How to check | Fix if violated |
|---|---|---|
| Independence | Study design | Survey weights, clustered SEs |
| Linearity | Scatter plot Y vs X | Transform variables (e.g., log) |
| Normal errors | Histogram of residuals | Usually OK in large samples |
| Homoskedasticity | Residuals vs. fitted plot | Robust standard errors (`vcov = "HC1"`) |

:::

## Why This Matters for Your Research Paper {.smaller}

- **HW #9** asks you to run an OLS regression on your research data — today's tools are exactly what you need
- Your **revised proposal (due 4/16)** should now describe which regression model you plan to use
- Your **final paper** will likely use OLS regression as its central analysis

::: {.callout-note icon=false}
## What HW #9 asks — and what you now know how to do:

1. **Identify your outcome and key predictor** → Part 1 today
2. **Run `lm()` and display results with `modelsummary()`** → Part 2 today
3. **Set a baseline if using a categorical predictor** → Part 3 today
4. **Interpret the coefficient magnitude, p-value, and R²** → Part 2 today
5. **Visualize** the relationship with a scatter + regression line (continuous IV) or coefficient plot → Part 2 today
:::

::: {.callout-tip icon=false}
## One thing to add to your thinking for hw9:
When interpreting your results, distinguish **statistical significance** (is $p < 0.05$?) from **substantive significance** (does the size of the coefficient actually matter in the real world?). Both belong in your write-up.
:::

## Questions?

## Assignments {.smaller}

**Weekly Assignment #9** — <span class="highlight">Due Thursday, April 9</span>

Using your research dataset:

1. State a research question with a **continuous outcome variable** and at least one **key predictor**
2. Write your hypothesis (direction of expected relationship and justification)
3. Run OLS regression with `lm()` and display results with `modelsummary()`
4. **If using a categorical predictor**: use `fct_relevel(var, "baseline_category")` to set a baseline, and state what it is
5. Interpret the estimated coefficient(s): direction, magnitude, and what they mean
6. Report and interpret **R²**
7. Visualize: scatter plot with regression line (continuous IV) or coefficient plot
8. Discuss both **statistical** and **substantive** significance

**Also due:**

- **HW #8** — <span class="highlight">Due Thursday, April 2</span> (hypothesis testing)
- **Revised Proposal with Outline** — <span class="highlight">Due Thursday, April 16</span>

**In-class lab today:**

- Practice running OLS on your own dataset and interpreting the output
