---
title: "Week 1"
subtitle: "Sociology 106: Quantitative Sociological Methods"
date: 01/20/2026
date-format: long
format:
  revealjs: 
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    css: slides.css
    footer: '[GitHub course repo](https://www.kaseyzapatka.com/soc106)'
---

## Agenda 

Introductions, welcome, and a few remarks about the course

Discussion of each of today’s learning goals:

- Have a sense of what the course is, what it is not, and what we'll cover this semester
- Understand structure and requirements of the course
- Download and install `R` and Positron

Install `R` and Positron -- <span class="highlight">should be done by class next week</span>

## Introductions

[Kasey Zapatka](http://kaseyzapatka.com), **Postdoctoral Researcher** 

- PhD, Sociology, CUNY Graduate Center
- Urban inequality with a focus on housing, affordability, and neighborhood change
  - Dissertation: how gentrification influences housing affordability and traced housing affordability over life course 
  - R/python, Machine Learning, MLM, causal inference, spatial econometrics

## Introductions

**Teaching**

- [Computational Social Science](https://github.com/kaseyzapatka/Computational-Social-Science-Training-Program) - course covering machine learning, text analysis, and causal inference

**Current Research**

- Developing a longitudinal database of evictions 
- Evaluating impact of Emergency Rental Assistance and tenant protections on evictions during COVID-19 pandemic 

## Now your turn

- Name
- Year
- Experience with `R` (no right answer)
- Mac or Windows user? (there's a right answer)
- Why are you taking this course?
- What are you most looking forward to in this course?


## Sociology 106 is…

...an intermediate undergraduate social science research methods course emphasizes the motivation, computation, and interpretation of statistical tests.

**Course Content**

- Statistical tests for one or two continuous or categorical variables
- Interpretation of linear and logistic regression
- Introduction to machine learning framework in the last week
- Data management and analysis using `R` statistical programming language

**Skills Developed**

Students will gain practical experience in `R` for conducting statistical analyses and managing data.

## Sociology 106 is not…

...about math. Statistics is about adjudicating between rival explanations of phenomena about a population, using data from that population.

- This course will be taught with more of a focus on the math / statistical theory or with more of a focus on sociological applications

- We will be more applied, still with some math (and some equations)

...about the “one true way to answer sociological questions” *(where quals at?)*.

## After Sociology 106, you will:

1. Understand the logic of statistical inference
2. Identify appropriate statistical tests for different types of data
3. Visualize data and produce descriptive statistics and simple statistical tests using `R`
4. Interpret and communicate statistical results and discuss their relevance in the context of a particular research question


## So, what is a statistical question?

> In past semesters, has Sociology 106 been more appealing to male or female students?

- One possibility: gender doesn’t make a difference!
- Another possibility: male students are more likely to have taken Sociology 106 in the past

## Could look at descriptive statistics

Some descriptive statistics…(from a previous semester!)

```{r}


# load libraries
# ---------
library(tidyverse)
library(here)

# data
# ----------
students_summary <- data.frame(
  gender = c("male", "female"),
  count = c(8, 3)
)

# plot
# ----------
students_summary |> 
  ggplot(aes(x = gender, y = count)) +
  geom_col(fill = "#0d467cff", width = 0.6) +
  scale_y_continuous(breaks = 0:9, limits = c(0, 9)) +
  labs(
    x = NULL,
    y = "Number of students",
    caption = "Percentage of male students = 72.2%"
  ) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.caption = element_text(size = 14, hjust = 0.5, margin = margin(t = 10))
  )

```


## The probability model

```{r}

# Create data for the binomial distribution
df <- tibble(
  male_students = 0:11,
  probability = dbinom(male_students, size = 11, prob = 0.5)
)

# Plot
ggplot(df, aes(x = male_students, y = probability)) +
  geom_col(fill = "steelblue", width = 0.6) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 0.25)
  ) +
  scale_x_continuous(breaks = 0:11) +
  labs(
    title = "Assuming P(student is male) = 50%, for a class of 11 students:",
    x = "Number of male students",
    y = "Probability of outcome"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.minor = element_blank()
  )

```

## One example: 

> In past semesters, has Sociology 106 been more appealing to male or female students?

One possibility: gender doesn’t make a difference!

- Null hypothesis: P(student is male) = 50%

Another possibility: male students are more likely to major in Statistics

- Alternative hypothesis: P(student is male) > 50%



## A statistical test

```{r}


# Parameters
n <- 11
p <- 0.5

# Create probability distribution
df <- tibble(
  male_students = 0:n,
  prob = dbinom(male_students, size = n, prob = p)
) %>%
  mutate(
    highlight = male_students >= 8
  )

# Compute tail probability
tail_prob <- sum(df$prob[df$male_students >= 8])

# Plot
ggplot(df, aes(x = male_students, y = prob, fill = highlight)) +
  geom_col(width = 0.6) +
  scale_fill_manual(
    values = c("FALSE" = "#5b9bd5", "TRUE" = "#c00000"),
    guide = "none"
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 0.26)
  ) +
  scale_x_continuous(breaks = 0:11) +
  labs(
    x = "Number of male students",
    y = "Probability of outcome"
  ) +
  annotate(
    "text",
    x = 8.6,
    y = 0.115,
    label = paste0(
      "P(# male students ≥ 8) = ",
      scales::percent(tail_prob, accuracy = 0.1)
    ),
    color = "#c00000",
    size = 5,
    hjust = 0.2
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank()
  )


```


## The same test, but in R

```{r echo=TRUE}

binom.test(c(8,3), p = 0.5, alternative="greater")

```

## By the end of this course…

You will be able to test:

- Theories involving one categorical or continuous variable.
  - Ex: gender is often measured as a categorical variable (male / female / other)
  - Ex: income is often measured as a continuous variable (the number of dollars one earns is a real number)
- Theories involving how one (or more) binary or continuous variable affects another binary or continuous variable.

## Questions?

## Course Expectations{.smaller} 

- **Active learning** - Lecture is for you so interrupt to ask questions if you have them
- **Safe and productive learning space** for researchers using these methods and supporting each other
- **Expose you to a lot** of coding and technical things but we can do it together
    - We're only learning the basics and that's all I expect from you
    - Coding in `R` can be intimidating, but promise it's worth learning how use
    - Positron is cutting edge IDE (most of grad students don't use), but it's the future of social science research

## Open week{.smaller} 

Right now I have an open week at the end and I want to fill it with something that is useful to you all. Material will likely not be included on final exam. 

Some options:

- brief intro to **machine learning**
- survey of more **advanced methods** (e.g., fixed effects, MLM, spatial regression, Poisson/Negative Binomial, etc.)
- open to class suggestions of what would be useful to you all

Let me know and we can take a vote in week 3 

## Course Navigation{.smaller} 


::::: columns
::: {.column width="50%"}

[bCourse](https://bcourses.berkeley.edu/courses/1551992) is our bible

- home 
- announcements
- assignments
- pages

:::
::: {.column width="50%"}

[GitHub course repo](https://www.kaseyzapatka.com/soc106/)

- slides (each week)
- syllabus (pdf on bCourse)

:::
:::::



## Required Course Readings{.smaller}

1. **David Lane's Online Statistics** (<http://onlinestatbook.com/Online_Statistics_Education.pdf>). 

2. **Hadley Wickham's R for Data Science 2nd Edition** (<https://r4ds.hadley.nz>), which is also free online.

3. We'll read two journal articles to see how social scientists use regression analysis in practice. 

  - **Thompson, M.S., & Keith, V. M.** (2001). The Blacker the Berry: Gender, Skin Tone, Self-Esteem, and Self-Efficacy. Gender & Society, 15(3), 336-357. (<https://doi.org/10.1177/089124301015003002>) 
  
  - **Freeman, L., & Braconi, F.** (2004). Gentrification and Displacement New York City in the 1990s. Journal of the American Planning Association, 70(1), 39–52. (<https://doi.org/10.1080/01944360408976337>)


## Required Course Materials

Access to a **laptop** where you can download: 

1. **`R`** (<http://cran.rstudio.com>), an open-source programming language

2. **Positron** (<https://positron.posit.co/download.html>), a free program that makes working in `R` much easier and is at the cutting edge of data science right now. 

We’ll start installation of `R` and Positron at the end of class--<span class="highlight">this must be done by class next week</span>


## Course Elements

- Attendance and participation (10%)
- Weekly homework assignments  (30%)
- Research paper (40%)
- Final exam (20%)

## Class Format

**First half** of class (give or take) will be a lecture where I go over new statistical concepts and show you how to implement these statistical concepts in `R`

<br>

::: {.fragment .fade-up}
**Second half** of class, either you will have time: 

  1. to start hw assignments that practice implementing these concepts in `R`
  2. we'll have extra time for questions/followups
  3. we'll discuss the research papers a bit more in-depth
::: 

## Typical Weekly Assignment{.smaller}

- Anywhere from **3-10 problems** with brief analysis and write-up in `R` and quarto
- Be sure to show your work -- **writing out conclusions** and interpreting results as necessary
- With couple of exceptions, you'll **work with your own dataset** to apply methods introduced in lecture
- Ideally the same dataset from week to week (use the same dataset for your research paper)
- For the first assignment, I will provide a dataset on bCourses
- Think carefully about the method, the types of questions it can answer, and the types of variables that can be used in the method  before choosing variables from your dataset
- Write up using **quarto (.qmd file)** template provided on bCourses to **submit to bCourse**


## Research Paper{.smaller}

You will develop and present a research question of your choice, address it using statistical techniques from the course that you apply to data in `R`, and write a paper summarizing your findings

- You can *(and should)* use your weekly assignments to work on your research question

- Several milestones throughout the semester will help you progress towards the final paper
  - Paper proposal (5%)
  - Annotated bibliography (5%)
  - Revised paper proposal with outline (5%)
  - In-class presentation (5%)  -- 7-10 mins
  - Final paper (20%)


## Keys to Success{.smaller}

<span class="highlight">Material is cumulative, so it is critical to keep up</span>

- Please ask questions during lecture!
- If you find yourself falling behind, seek help immediately from me during office hours

<span class="highlight">Learning statistics requires thinking through how to solve problems</span>

- This is what the weekly assignments are for; you should not expect to fully understand the material until after you have completed the assignment
- Feel free to work on assignments in groups, though what you turn in must be your own work.

## Keys to Success{.smaller}

<span class="highlight">Learning statistics is like learning a language</span>

- The material in this course can be challenging / counterintuitive if you haven’t seen it before
- It is important not to be intimidated by new terms or the use of letters to represent quantities or variables
- Review your algebra skills if necessary

<span class="highlight">Lecture slides will be made available, but are not a substitute for careful note taking</span>


## Office Hours{.smaller}

I can help in office hours with questions about concepts from lecture or about coding in `R`

- **Tuesdays**, right before 11:30 AM-1:00 PM, TBA
- **Tuesdays**, right after class, until 4:30 PM, TBA
    - Sign up at <https://www.wejoinin.com/kaseyzapatka@berkeley.edu>
- If you can’t make either, e-mail me and we’ll find a time

Outside of office hours, your best resources are your fellow classmates!

- If something is really unclear to you and others, e-mail me and we can go over it in next week’s lecture

## Online and Generative AI Policy{.smaller}

- <span class="highlight">PLEASE USE THEM!</span>--AI tools (e.g., ChatGPT), Google, and online resources (e.g., [StackOverflow](https://stackoverflow.com/questions)  or [StackExchange](https://stackexchange.com))
    - <span class="highlight">DO</span> use these tools to better understand concepts, debug code, suggest libraries or packages, or offer sample code to guide learning. 
    - <span class="highlight">DO NOT</span> copy or paraphrase AI-generated content in any assignment, as this counts as plagiarism
- <span class="highlight">The goal</span> is to use them to support your learning and problem-solving, not to replace it. 

> If you are going to spend the time taking this course, you should learn as much as you can.

## Questions?


## Getting started with `R` and Positron

Download and install `R`:[^Rdownload]  
  <https://cran.rstudio.com>


Download and install Positron[^Positrondownload]  
	<https://positron.posit.co/download.html>

[^Rdownload]: Mac users need to check whether they have a Apple Silicon (M1/M2) or Intel processor and download the appropriate version of `R`for Mac:   
1. Click the Apple menu () in the top-left corner of your screen.  
2. Select **About This Mac**.  
3. Look at the **Processor** or **Chip** line:  
   - If it says **Apple M1**, **M2**, or similar → you have Apple Silicon.  
   - If it says **Intel Core i5**, **i7**, etc. → you have an Intel Mac.

[^Positrondownload]: You can just download the "Universal" option, but know it is slightly larger file and might not be quite as optimized. 