---
title: "Weekly Assignment #2"
author: "Your Name Here"  # Replace with your name
date: "Due Thursday, February 12 at 11:59 PM"
format: 
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

::: callout-important
## Submission Instructions

1.  Download this `hw2.qmd` file from bCourse and download your own data separately.
2.  Save both files on your computer. Make sure to save your data in your data folder and `hw2.qmd` in your assignments folder.
3.  Follow the instructions below to complete the assignment. A sample code chunk has been provided but you might need more.
4.  When you are done, click the "Render" button at the top of Positron to generate an .html file (Note that it might not necessarily save the .html file, but if you can preview it, that means it's working).
5.  Submit both `hw2.qmd` and `hw2.html` files to bCourse before the deadline
6.  Make sure your name appears in the "author" field in the YAML header above
:::

You're folder structure should look something like this after you upload `hw2.qmd`:

```         
soc106/
├── _quarto.yml
├── data/
│   └── attain.csv
│   └── [[your_data]].csv
├── assignments/
│   └── hw0.qmd
│   └── hw1.qmd
│   └── hw2.qmd
└── labs/
    └── lab1.qmd
```

## Instructions

Using a dataset of your choice, choose a research question and construct a plot that helps answer it. Make sure to answer the questions below and that your document renders without errors before submitting.

```{r}
# Clear environment
# ---------
rm(list = ls())

# install libraries -- only install new libraries if needed
# ---------
# install.packages("tidyverse")

# load libraries
# ---------
library(tidyverse)
library(here)
# add more libraries here if you need them!


```

You'll want to import your data. I've commented out some code that you can use to get started, but you'll need to update it to make it work with your data. If your data is in the format of `.csv` then you can use `read_csv` to load it. If not, check out [this reference website](https://libguides.chapman.edu/R/import).

```{r}

# Load data
#data <- read_csv(here("data", "[[your_data]].csv"))

# Your code here

```

## Research question

**What is your research question that you want to answer with this plot? (Remember a research question should end with a "?")**

*Insert your answer here*

## Prediction and brief justification

**State what you expect to find and why. The justification can be based on theory or previous research. Even better (where possible) is to set up competing predictions based on competing theoretical premises: e.g., “If Marx is right, we should find P; but if Weber is right, then we should find Q.**

*Insert your answer here*

## Data

**Describe the data in one or two sentences. Then state in one or two sentences why you’re using this data instead of different data.**

*Insert your answer here*

## Analytic strategy

**Describe the type of plot you are going to use to analyze the data, including the names of plot and the general R commands. Why are you using this type of plot instead of another? A few sentences is fine.**

*Insert your answer here*

## Code

**Build your plot here. If you need to add more code chunks, feel free to add more (using the drop down menu from Insert in Visual mode or "Insert Code Cell" option in Souce mode), but you can probably do it all in one code chunk.**

```{r}

# Your code here

```

## Results

**Present the results of the analysis here. What did you learn from the figure? This is not the place for extended commentary, just the basic results of your analysis (e.g., X category has a lot of observations, there seems to be a relationship between X and Y). Should only be a few sentences.**

*Insert your answer here*

## Discussion of results

**Here’s where you talk about how your predictions went and what the findings mean about the answer to your research question. (e.g., Were you wrong, were you right? Why might that have been?) One to three sentences will do.**

*Insert your answer here*

## Limitations and alternative explanations

**This is where you talk about what you might have done differently or better or what other people should do next time. All research is flawed in dozens of ways, so for this assignment please talk about one limitation of your data, one limitation of your analysis, and one possible alternative explanation for your findings. What else might you need to better address this question? Two to three sentences is fine.**

*Insert your answer here*

::: callout-tip
## Before You Submit

-   [ ] Your name is in the author field at the top of this document
-   [ ] All code chunks run without errors
-   [ ] You've answered all written response questions
-   [ ] You've clicked "Render" and successfully generated an .html file
-   [ ] You're submitting both `hw2.qmd` and `hw2.html` files to bCourse
:::