---
title: "Weekly Assignment #10"
author: "Your Name Here"  # Replace with your name
date: 04/23/2026
date-format: long
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

::: callout-important
## Submission Instructions

1.  Download this `hw10.qmd` file from bCourse and use the dataset for your final research paper.
2.  Save both files on your computer. Make sure to save your data in your data folder and `hw10.qmd` in your assignments folder.
3.  Follow the Weekly Assignment Analysis Template structure below to complete the assignment.
4.  When you are done, click the "Render" button at the top of Positron to generate an .html file.
5.  Submit the `hw10.qmd` file to bCourse before the deadline.
6.  Make sure your name appears in the "author" field in the YAML header above.
:::

# Instructions

Develop a hypothesis about how one (or more) independent variables (in the dataset you will use in your final paper) may be related to a **binary** dependent variable of interest (Example: I predict X is positively related to Y). Then, test that hypothesis using a logistic regression, and write up your analysis using the Weekly Assignment Analysis Template structure below.

**Important Notes:**

-   If you use a categorical independent variable, make sure to set a baseline category, and write down in your assignment what the baseline category is!
-   In writing up your results, first report whether the relationship is significant in the predicted direction.
    -   Interpret the coefficients for the key variable. They should be odds ratios (not logged odds) and it's best to report the percentage change in the odds — computed as $(OR - 1) \times 100$ — for each coefficient.
    -   Report predicted probabilities at two or more values of the independent variable using a margins plot.

## Setup

```{r}
#| label: setup
#| warning: false
#| message: false

# Clear environment
rm(list = ls())

# Load libraries
library(tidyverse)
library(here)
library(modelsummary)
library(ggeffects)

# Turn off scientific notation
options(scipen = 999)
```

------------------------------------------------------------------------

## Question

**Write your research question here as a single sentence ending in a question mark. This should involve the relationship between your independent variable(s) and a binary dependent variable.**

*Your research question:*

------------------------------------------------------------------------

## Prediction(s) and Brief Justification

**State what you expect to find and why. Be specific about the direction of relationships you expect. The justification can be based on theory or previous research. Even better (where possible) is to set up competing predictions based on competing theoretical premises.**

*Your prediction and justification (3-5 sentences):*

------------------------------------------------------------------------

## Data

**Describe the data in one or two sentences. Then state in one or two sentences why you're using this data instead of different data. Be sure to describe your binary dependent variable and independent variable(s).**

*Your data description:*

------------------------------------------------------------------------

## Analytic Strategy

**Describe the exact procedure you are going to use to analyze the data, including:**

-   *Your binary dependent variable (Y) and independent variable(s) (X)*
-   *If using categorical variables, what the baseline category is*
-   *What your regression equation looks like*
-   *What your null hypothesis (H₀) and alternative hypothesis (H₁) are*
-   *What significance level (α) you will use*
-   *What specific R functions you will use*
-   *Why you are using logistic regression instead of OLS*

*This should be 3-6 sentences.*

*Your analytic strategy:*

------------------------------------------------------------------------

## Code and Analysis

**Load your data and conduct your analysis below. Show all relevant code.**

```{r}
# Load your data
# Drop missing values on all variables that enter the model (do this before glm())
# Create your binary (0/1) outcome variable
# If using a categorical predictor, set an explicit baseline with fct_relevel()

# your code here
```

```{r}
#| message: false
#| warning: false

# Visualize the proportion of Y = 1 across your key predictor
# (bar chart by group for categorical; binned bar chart for continuous)

# your code here
```

```{r}
# Fit the logistic regression model using glm(..., family = binomial)
# Use summary() to see the raw output

# your code here
```

```{r}
# Display a clean results table using modelsummary()
# Use vcov = "HC1" for robust SEs and exponentiate = TRUE for odds ratios
# Note: the reference line for significance is OR = 1, not 0

# your code here
```

```{r}
#| fig-height: 4

# Visualize the odds ratios using modelplot()
# Use exponentiate = TRUE and add geom_vline(xintercept = 1, ...)

# your code here
```

```{r}
#| message: false

# Margins plot: use ggpredict() to show predicted probabilities across your key predictor
# You can also plot by a grouping variable using terms = c("X1", "X_group")

# your code here
```

------------------------------------------------------------------------

## Results

**Present the results of the analysis here. This is not the place for extended commentary, just the basic results of your analysis. Include:**

1.  *Whether the relationship is significant in the predicted direction*
2.  *Your interpretation of the odds ratio for each independent variable and the percentage change in odds:* $(OR - 1) \times 100$
3.  *The predicted probabilities at two or more values of the independent variable (from your margins plot)*
4.  *Whether you reject or fail to reject your null hypothesis*

**This should be 5-8 sentences.**

*Your results:*

------------------------------------------------------------------------

## Discussion of Results

**Here's where you talk about how your predictions went and what the findings mean about the answer to your research question. Discuss both the statistical significance and the practical/substantive significance of your findings. One to three sentences will do.**

*Your discussion:*

------------------------------------------------------------------------

## Limitations and Alternative Explanations

**This is where you talk about what you might have done differently or better or what other people should do next time. All research is flawed in dozens of ways, so for this assignment please talk about one limitation of your data, one limitation of your analysis, and one possible alternative explanation for your findings. Three to five sentences is fine.**

*Your limitations and alternative explanations:*

------------------------------------------------------------------------

::: callout-tip
## Before You Submit

-   [ ] Your name is in the author field at the top of this document
-   [ ] You've clearly identified your binary dependent variable and independent variable(s)
-   [ ] If using categorical variables, you've specified the baseline category
-   [ ] You've reported odds ratios and the percentage change in odds for each coefficient
-   [ ] You've produced and included a margins plot showing predicted probabilities
-   [ ] You've stated your null and alternative hypotheses
-   [ ] All code chunks run without errors
-   [ ] You've answered all sections of the analysis template
-   [ ] You've discussed both statistical and substantive significance
-   [ ] You've clicked "Render" and successfully generated an HTML file
-   [ ] You're submitting the `.qmd` file to bCourse
:::