---
title: "Weekly Assignment #10"
author: "Your Name Here"  # Replace with your name
date: 04/23/2026
date-format: long
format: 
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

::: callout-important
## Submission Instructions

1.  Download this `hw10.qmd` file from bCourse and use the dataset for your final research paper.
2.  Save both files on your computer. Make sure to save your data in your data folder and `hw10.qmd` in your assignments folder.
3.  Follow the Weekly Assignment Analysis Template structure below to complete the assignment.
4.  When you are done, click the "Render" button at the top of Positron to generate an .html file.
5.  Submit the `hw10.qmd` file to bCourse before the deadline.
6.  Make sure your name appears in the "author" field in the YAML header above.
:::

# Instructions

Develop a hypothesis about how one (or more) independent variables (in the dataset you will use in your final paper) may be related to a binary dependent variable of interest. (Example: I predict X is positively related to Y). Then, test that hypothesis using a logistic regression, and write up your analysis using the Weekly Assignment Analysis Template structure below.

**Important Notes:**

-   If you use a categorical independent variable, make sure to set a baseline category, and write down in your assignment what the baseline category is!
-   In writing up your results, first report whether the relationship is significant in the predicted direction. Next, report the percentage change in the odds given by the coefficient. Finally, report the predicted probability at two or more values of the independent variable.

## Setup

```{r}
#| label: setup
#| warning: false
#| message: false

# Clear environment
rm(list = ls())

# Load libraries
library(tidyverse)
library(here)
library(modelsummary)
# Add more libraries here if you need them!

```

------------------------------------------------------------------------

## Question

**Write your research question here as a single sentence ending in a question mark. This should involve the relationship between your independent variable(s) and a binary dependent variable.**

*Your research question:*

------------------------------------------------------------------------

## Prediction(s) and Brief Justification

**State what you expect to find and why. Be specific about the direction of relationships you expect. The justification can be based on theory or previous research. Even better (where possible) is to set up competing predictions based on competing theoretical premises.*
*

*Your prediction and justification (3-5 sentences):*

------------------------------------------------------------------------

## Data

**Describe the data in one or two sentences. Then state in one or two sentences why you're using this data instead of different data. Be sure to describe your binary dependent variable and independent variable(s).**


*Your data description:*

------------------------------------------------------------------------

## Analytic Strategy

**Describe the exact procedure you are going to use to analyze the data, including:**

-   *Your binary dependent variable (Y) and independent variable(s) (X)*
-   *If using categorical variables, what the baseline category is*
-   *What your regression equation looks like*
-   *What your null hypothesis (H₀) and alternative hypothesis (H₁) are*
-   *What significance level (α) you will use*
-   *What specific R functions you will use*
-   *Be sure to explain why you are using logitistic regression instead of linear regression*

*This should be 3-6 sentences, but no longer than 1/3 of a page.*


*Your analytic strategy:*

------------------------------------------------------------------------

## Code and Analysis

**Load your data and conduct your analysis below. Show all relevant code.**

### Load your data

```{r}

# Load your data
# Your code here

```


### Data cleaning 

```{r}

# Any data cleaning or variable creation needed
# Create binary dependent variable if necessary
# If using categorical variables, create factors and set baseline categories
# Your code here

```

### Any descriptives

```{r}

# Show descriptive statistics for your key variables
# Your code here

```

### Run your regression

```{r}

# Run your logistic regression
# Your code here
# Example: model <- glm(outcome ~ x1 + x2, data = mydata, family = binomial(link = "logit"))

```

### Display your results 

```{r}
 
# Display a clean summary of your regression results
# You can use summary(), modelsummary(), or other formatting functions. 
# Your code here

```

### Odds ratio

```{r}

# Calculate and display odds ratios (exponentiated coefficients)
# Your code here
# Example: exp(coef(model))
# Or: exp(confint(model))

```

### Predicted probabilities

```{r}

# Calculate predicted probabilities at different values of your independent variable(s)
# Your code here
# You may want to create a new data frame with specific values and use predict()

```

------------------------------------------------------------------------

## Results

**Present the results of the analysis here. This is not the place for extended commentary, just the basic results of your analysis. Include:**

1.  *Whether the relationship is significant in the predicted direction*
2.  *The log-odds coefficients for each independent variable*
3.  *The percentage change in odds given by each coefficient*
4.  *The predicted probabilities at two or more values of the independent variable*
5.  *Whether you reject or fail to reject your null hypothesis*

**This should be 5-8 sentences.**

*Your results:*

------------------------------------------------------------------------

## Discussion of Results

**Here's where you talk about how your predictions went and what the findings mean about the answer to your research question. Discuss both the statistical significance and the practical/substantive significance of your findings. One to three sentences will do.**


*Your discussion:*

------------------------------------------------------------------------

## Limitations and Alternative Explanations

**This is where you talk about what you might have done differently or better or what other people should do next time. All research is flawed in dozens of ways, so for this assignment please talk about one limitation of your data, one limitation of your analysis, and one possible alternative explanation for your findings. Three to five sentences is fine.**

*Your limitations and alternative explanations:*

------------------------------------------------------------------------

::: callout-tip
## Before You Submit

-   [ ] Your name is in the author field at the top of this document
-   [ ] You've clearly identified your binary dependent variable and independent variable(s)
-   [ ] If using categorical variables, you've specified the baseline category
-   [ ] You've reported log-odds coefficients AND odds ratios (percentage change in odds)
-   [ ] You've calculated and reported predicted probabilities at different values
-   [ ] You've stated your null and alternative hypotheses
-   [ ] All code chunks run without errors
-   [ ] You've answered all sections of the analysis template
-   [ ] You've discussed both statistical and substantive significance
-   [ ] You've included your complete R output in the appendix
-   [ ] You've clicked "Render" and successfully generated an HTML file
-   [ ] You're submitting the .qmd file to bCourse
:::