---
title: "Weekly Assignment #5"
author: "Your Name Here"  # Replace with your name
date: 03/05/2026
date-format: long
format: 
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

::: callout-important
## Submission Instructions

1.  Download this `hw5.qmd` file from bCourse.
2.  Save the file in your assignments folder.
3.  Follow the instructions below to complete the assignment. Show your work for all calculations!
4.  When you are done, click the "Render" button at the top of Positron to generate an .html file. If you can preview it on the left pane, that means it's working.
5.  Submit both the `hw5.qmd` and the `hw5.html` files to bCourse before the deadline
6.  Make sure your name appears in the "author" field in the YAML header above
:::

# Instructions

Complete all questions below, writing your code in the code chunks provided. Replace `# Your code here` with your actual code. For questions requiring hand calculations, show your work in the text sections. Make sure your document renders without errors before submitting.

## Setup

```{r}
#| label: setup
#| warning: false
#| message: false

# Clear environment
# ---------
rm(list = ls())

# install libraries -- shouldn't need for this assignment
# ---------
# install.packages("tidyverse")

# load libraries
# ---------
library(tidyverse)
library(here)
```

------------------------------------------------------------------------

## Question 1: Free Throw Probability

**Lisa makes free throws 75% of the time, and shoots eight free throws at the end of practice.**

**a) In words, describe what you would need to calculate to find the probability that Lisa makes at least seven free throws. What are the values of n and π, and which outcomes do you need to add up?**

*Your answer here: `We need P(X ≥ 7) where X ~ Binomial(n = 8, π = 0.75). "At least seven" means either exactly 7 or exactly 8 made free throws, so we calculate P(X = 7) + P(X = 8) using the binomial formula (or dbinom()) with n = 8 and π = 0.75.`*

**b) In R: what is the probability that she will make between four and six free throws?**

```{r}

# Your code here
dbinom(4, size = 8, prob = .75) + dbinom(5, size = 8, prob = .75) + dbinom(6, size = 8, prob = .75) # .6056213
# OR
pbinom(6, size = 8, prob = .75) - pbinom(3, size = 8, prob = .75) # .6056213


```

*Your answer here: `0.6056213`*

------------------------------------------------------------------------

## Question 2: Exam Performance

**Javier takes a 50 question exam. In practice exams, he has answered each question correctly 80% of the time, and his likelihood of answering one question correctly is independent of whether he answers any other question correctly.**

**a) What is the mean number of questions he will get correct on the exam? What is the standard deviation of the number of questions he will get correct?**

*Show your work here:*

```{r}

# Your code here
# mean 
50*.8 # 40

# SD
sqrt(50*.8*.2) # 2.828...

```

*Your answer here: `mean: 40, SD: 2.828`*

**b) A student needs to get 70% of questions correct to pass the exam. What is the probability that he will pass the exam? (use R to answer)**

```{r}

# Your code here
1 - pbinom(34, 50, .8) 

```

*Your answer here: `0.9691966`*

------------------------------------------------------------------------

## Question 3: Height and Z-scores

**Andre is selected from a population where height is distributed normally with mean 69 inches and standard deviation 3 inches. The individual is measured with a height of 75 inches.**

**a) Calculate the z-score of Andre's height.**

*Show your work here:*

```{r}

# Your code here
(75-69)/3 # 2

```

*Your answer here: 2*

**b) Just from the z-score, can you tell whether Andre's height is above the 95th percentile of height? Why or why not? How about the 97th percentile of height? Why or why not?**

*Your answer here: `Yes. Andre's z-score is 2, which places him about 2 standard deviations above the mean. In a normal distribution, z = 2 corresponds to about the 97.7th percentile (pnorm(2) ≈ 0.977). So his height is above both the 95th percentile and the 97th percentile.`*

------------------------------------------------------------------------

## Question 4: Vehicle Speeds on I-5

**Assume the speed of vehicles along a stretch of I-5 has an approximately normal distribution with a mean of 71 mph and a standard deviation of 8 mph.**

**a) The current speed limit is 65 mph. What is the proportion of vehicles less than or equal to the speed limit?**

```{r}

# Your code here
pnorm(65, 71, 8) 


```

*Your answer here: `0.2266274`*

**b) What proportion of the vehicles would be going less than or equal to 50 mph?**

```{r}

# Your code here
pnorm(50, 71, 8) 

```

*Your answer here: `0.00433`*

**c) A new speed limit will be initiated such that approximately 10% of vehicles will be over the speed limit. What is the new speed limit based on this criterion?**

```{r}

# Your code here
qnorm(.9, 71, 8) 

```

*Your answer here: `81.25241`*

------------------------------------------------------------------------

## Question 5: NCAA SAT Requirements

**The NCAA requires Division I athletes to score over 820 on the combined math and verbal SAT exam to compete in their first college year. The SAT scores of 2003 were approximately normal with mean 1026 and standard deviation 209.**

**a) What proportion of all students would be NCAA qualifiers (SAT \> 820)?**

```{r}

# Your code here
1 - pnorm(820, 1026, 209) 

```

*Your answer here: `0.837`*

**b) The NCAA defines a "partial qualifier" eligible to practice and receive an athletic scholarship, but not to compete, as a combined SAT score over 720 but less than or equal to 820. What proportion of all students who take the SAT would be partial qualifiers?**

```{r}

# Your code here
pnorm(820, 1026, 209) - pnorm(720, 1026, 209) 

```

*Your answer here: `0.09057`*

::: callout-tip
## Before You Submit

-   [ ] Your name is in the author field at the top of this document
-   [ ] All code chunks run without errors (or you've indicated hand calculations)
-   [ ] You've shown your work for all probability calculations
-   [ ] You've answered all written response questions
-   [ ] You've clicked "Render" and successfully generated an HTML file
-   [ ] You're submitting both `hw5.qmd` and `hw5.html` files to bCourse
:::