---
title: "Weekly Assignment #6 — Key"
author: "Your Name Here"  # Replace with your name
date: 03/12/2026
date-format: long
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

::: callout-important
## Submission Instructions

1.  Download this `hw6.qmd` file from bCourse.
2.  Save the file in your assignments folder.
3.  Follow the instructions below to complete the assignment. Show your work for all calculations where necessary!
4.  When you are done, click the "Render" button at the top of Positron to generate an .html file.
5.  Submit both `hw6.qmd` and `hw6.html`file to bCourse before the deadline.
6.  Make sure your name appears in the "author" field in the YAML header above.
:::

# Instructions

Complete all questions below, writing your code in the code chunks provided. Replace `# Your code here` with your actual code. For questions requiring hand calculations, show your work in the text sections. Make sure your document renders without errors before submitting.

## Setup

```{r}
#| label: setup
#| warning: false
#| message: false

# Clear environment
# ---------
rm(list = ls())

# load libraries
# ---------
library(tidyverse)
library(here)
```

------------------------------------------------------------------------

## Question 1: The Three Distributions

The University of California system reports that **22% of UC students** are first-generation college students (meaning neither parent attended a four-year college). A researcher is studying whether first-generation students are proportionally represented among recipients of a competitive academic fellowship. Over the past two years, **8 out of 80** fellowship recipients were first-generation students.

**a) In 1–2 sentences each, describe what the *population distribution*, the *sample (data) distribution*, and the *sampling distribution* each refer to in this scenario. No calculations needed.**

*Answer:*

-   **Population distribution**: the distribution of first-generation status among all UC students, where 22% are first-generation.
-   **Sample (data) distribution**: the distribution of first-generation status among the 80 fellowship recipients we actually observed (8 of 80, or 10%, were first-generation).
-   **Sampling distribution**: the distribution of $\hat{p}$ — the proportion of first-generation students — that we would observe if we repeatedly drew random samples of 80 recipients from the UC student population. It describes how much $\hat{p}$ would vary from sample to sample under random selection.

**b) Under random selection, what proportion of the 80 fellowship recipients would we *expect* to be first-generation students? Calculate the standard error of the sampling distribution of** $\hat{p}$ for $n = 80$. Show your work.

```{r}

# standard error
sqrt(0.22 * 0.78 / 80)

```

*Answer: Under random selection, we would expect* $\pi = 0.22$ of recipients (about 17–18 of 80) to be first-generation. The standard error is:

$$SE(\hat{p}) = \sqrt{\frac{0.22 \times 0.78}{80}} \approx 0.046$$

**c) The observed proportion is** $\hat{p} = 8/80 = 0.10$. How many standard errors is this from what we would expect under random selection? Also calculate the probability of observing a proportion this low or lower under random selection.

```{r}

# z-score
(0.10 - 0.22) / sqrt(0.22 * 0.78 / 80)

# probability of observing p-hat this low or lower
pnorm(0.10, mean = 0.22, sd = sqrt(0.22 * 0.78 / 80))

```

*Answer: The observed proportion is about **2.6 standard errors below** the expected value of 0.22. Under random selection, we would see a proportion this low or lower only about **0.5%** of the time.*

**d) Based on your calculation in (c), is the observed proportion consistent with random selection? What might this pattern suggest?**

*Answer: No — the observed proportion is 2.6 SEs below what we would expect if selection were random, and it would arise by chance less than 1% of the time. This pattern suggests that first-generation students may be systematically underrepresented in fellowship selection, possibly due to lower application rates, a lack of mentorship in the application process, or bias in the selection committee.*

------------------------------------------------------------------------

## Question 2: Sampling Designs

A team of researchers wants to study the relationship between neighborhood poverty rates and health outcomes across the United States. They plan to survey households in a sample of U.S. counties.

**a) Briefly explain how a *cluster sampling* design would work for this study. What is one advantage of cluster sampling compared to simple random sampling?**

*Answer: In cluster sampling, the researchers would first divide the country into clusters (counties), then randomly select a subset of those counties, and then randomly sample households within the selected counties. One advantage is cost and feasibility: it is much cheaper to survey households concentrated in a few counties than to send interviewers to a truly random nationwide sample of households.*

**b) The researchers decide to over-sample counties where poverty rates exceed 30% — drawing larger samples within those counties — so they have enough data to make comparisons within high-poverty areas. What type of sampling design is this, and what is the purpose of over-sampling this group?**

*Answer: This is **stratified random sampling**, where counties are divided into strata based on poverty rate. The purpose of over-sampling high-poverty counties is to ensure that there are enough respondents within that stratum to draw precise inferences about health outcomes in high-poverty areas. Without over-sampling, high-poverty counties (which are relatively rare) might produce too few observations to analyze reliably.*

**c) After collecting the data, the researchers find that households in rural counties were about half as likely to respond to the survey as households in urban counties. What is the appropriate statistical fix for this problem, and how does it work?**

*Answer: The appropriate fix is to apply **sampling weights**. Households that were less likely to respond (rural) receive higher weights so that they represent not just themselves but the similar rural households who did not respond. Weights are set inversely proportional to the probability of selection/response, making the weighted sample representative of the full population.*

------------------------------------------------------------------------

## Question 3: Jury Selection and Racial Representation

Jury selection is a constitutional right, and courts have found that systematic underrepresentation of racial groups violates the Equal Protection Clause. In a county in Texas, **28% of registered voters are Latino**. A civil rights organization examines records from the past year and finds that of **200 jury summons randomly issued**, only **42 went to Latino residents** ($\hat{p} = 42/200 = 0.21$).

**a) State the population parameter** $\pi$ and the sample statistic $\hat{p}$ for this scenario.

*Answer: The population parameter is* $\pi = 0.28$ (the proportion of registered voters who are Latino). The sample statistic is $\hat{p} = 0.21$ (the proportion of jury summons that went to Latino residents).

**b) Assuming jury summons are issued randomly, calculate the mean and standard error of the sampling distribution of** $\hat{p}$.

```{r}

# mean of sampling distribution
pi  <- 0.28
n   <- 200

mu_phat <- pi
mu_phat

# standard error
se_phat <- sqrt(pi * (1 - pi) / n)
se_phat

```

*Answer:*

$$\mu_{\hat{p}} = 0.28 \qquad SE(\hat{p}) = \sqrt{\frac{0.28 \times 0.72}{200}} \approx 0.032$$

**c) Calculate how many standard errors the observed proportion (**$\hat{p} = 0.21$) is from what we would expect under random selection. Then calculate the probability of observing a proportion this low or lower if jury summons were truly issued at random.

```{r}

# z-score (how many SEs below the expected value?)
p_hat_obs <- 0.21
z <- (p_hat_obs - mu_phat) / se_phat
z

# probability of observing p-hat this low or lower
pnorm(p_hat_obs, mean = mu_phat, sd = se_phat)

```

*Answer: The observed proportion is about **2.2 standard errors below** the expected value. Under random selection, the probability of observing* $\hat{p} \leq 0.21$ is approximately **1.4%** — a fairly rare outcome.

**d) Interpret your results. Is this pattern consistent with random jury selection? What might explain the gap between the expected and observed proportions?**

*Answer: A result this far from the expected value would occur only about 1–2% of the time if summons were truly random. This is strong evidence that Latino residents are being systematically underrepresented in this county's jury pool. Possible explanations include biased address lists used to generate summons (which may miss Latino neighborhoods), higher rates of disqualification among Latino residents (e.g., non-citizenship), or other administrative factors that produce non-random outcomes.*

------------------------------------------------------------------------

::: callout-tip
## Before You Submit

-   [ ] Your name is in the author field at the top of this document
-   [ ] All code chunks run without errors
-   [ ] You've answered all written response questions and shown your work
-   [ ] You've clicked "Render" and successfully generated an HTML file
-   [ ] You're submitting both the .qmd and .html file to bCourse
:::