---
title: "Weekly Assignment #6"
author: "Your Name Here"  # Replace with your name
date: 03/12/2026
date-format: long
format: 
  html:
    toc: true
    code-fold: false
    embed-resources: true
editor: visual
---

::: callout-important
## Submission Instructions

1.  Download this `hw6.qmd` file from bCourse.
2.  Save the file in your assignments folder.
3.  Follow the instructions below to complete the assignment. Show your work for all calculations where necessary!
4.  When you are done, click the "Render" button at the top of Positron to generate an .html file.
5.  Submit the `hw6.qmd` file to bCourse before the deadline.
6.  Make sure your name appears in the "author" field in the YAML header above.
:::

# Instructions

Complete all questions below, writing your code in the code chunks provided. Replace `# Your code here` with your actual code. Make sure your document renders without errors before submitting.

## Setup

```{r}
#| label: setup
#| warning: false
#| message: false

# Clear environment
# ---------
rm(list = ls())

# install libraries -- shouldn't need for this assignment
# ---------
# install.packages("tidyverse")

# load libraries
# ---------
library(tidyverse)
library(here)
```

------------------------------------------------------------------------

## Question 1: Sampling Distribution

**In a city, 70% of the people prefer Candidate A. Suppose 30 people from this city were sampled.**

**a) What is the mean of the sampling distribution of p?**

*Your answer here: `mean = .7`*

**b) What is the standard error of p?**

```{r}

# Your code here
sqrt(.7*.3/30) 

```

*Your answer here: `0.0836`*

**c) What is the probability that 80% or more of this sample will prefer Candidate A?**

```{r}

# Your code here
1 - pnorm(.8,.7,.0836666) 

# I also accepted using the binomial distribution here, which gives a slightly different answer. In larger sample sizes, these answers would be almost identical:
1 - pbinom(23,30,.7)

```

*Your answer here: `0.116 or 0.159`*

------------------------------------------------------------------------

## Question 2: Sample Size and Sampling Distribution

**In the population, the mean SAT score is 1000. Would you be more likely (or equally likely) to get a sample mean of 1200 if you randomly sampled 10 students or if you randomly sampled 30 students? Explain.**

*Your answer here: `You are more likely to get a mean of around 1,200 in a smaller sample because there is more variance in a smaller sample`*

------------------------------------------------------------------------

## Question 3: Normal Distribution and Sample Means

**Given a test that is normally distributed with a mean of 100 and a standard deviation of 12, find:**

**a) The probability that a single score drawn at random will be greater than 110**

```{r}

# Your code here
1 - pnorm(110,100,12) 

```

*Your answer here: `0..2023284`*

**b) The probability that a sample of 25 scores will have a mean greater than 105**

```{r}

# Your code here
1 - pnorm(105,100,12/5)

```

*Your answer here: `0..0186`*

**c) The probability that a sample of 64 scores will have a mean greater than 105**

```{r}

# Your code here
1 - pnorm(105,100,12/8) 

```

*Your answer here: `0.000429`*

**d) The probability that the mean of a sample of 16 scores will be either less than 95 or greater than 105**

```{r}

# Your code here
pnorm(95,100,12/4) + (1 - pnorm(105,100,12/4))

```

*Your answer here: `0.0956`*

------------------------------------------------------------------------

## Question 4: Comparing Two Schools

**The mean GPA for students in School A is 3.0; the mean GPA for students in School B is 2.8. The standard deviation in both schools is 0.25. The GPAs of both schools are normally distributed. If 9 students are randomly sampled from each school, what is the probability that:**

**a) The sample mean for School B will be greater than the sample mean for School A?**

```{r}

# Your code here
pnorm(0, .2, sqrt(.25^2/9 + .25^2/9)) 

```

*Your answer here: `0.0448`*

**b) The sample mean for School A will exceed that of School B by 0.5 or more?**

```{r}

# Your code here
1 - pnorm(.5, .2, sqrt(.25^2/9 + .25^2/9)) 

```

*Your answer here: `0.00545`*

------------------------------------------------------------------------

## Question 5: Freight Elevator Problem

**A large freight elevator can transport a maximum of 7600 pounds. Suppose a load of cargo containing 39 boxes must be transported via the elevator. Experience has shown that the weight of boxes of this type of cargo follows a distribution with mean µ = 200 pounds and standard deviation σ = 15 pounds. Based on this information, what is the probability that all 39 boxes can be safely loaded onto the freight elevator and transported?**

```{r}

# Your code here
pnorm(7600/39, 200, 15/sqrt(39)) 


```

*Your answer here: `To safely transport a sample of 39 boxes, their sampling mean must be below 7600/39. The probability of this is 0.01637879`*

------------------------------------------------------------------------

## Question 6: Wine Inventory Management

Suppose the number of bottles of wine a restaurant sells each week is normally distributed with mean 500 bottles and standard deviation of 100 bottles. The restaurant starts with 800 bottles, and has a scheduled delivery at the beginning of each week of 450 bottles.

**a) Find the probability that, after 10 weeks, the supply of wine will be below 200 bottles.**

```{r}

# Your code here
1 - pnorm(510,500,100/sqrt(10)) 

```

*Your answer here: `After 10 weeks, the restaurant will have an input of 800 + 450*10 = 5300 bottles of wine. If the restaurant is below 200 bottles of wine after 10 weeks, then the average weekly output must be greater than 5100/10 = 510. The probability of this is 0.3759148.`*

------------------------------------------------------------------------

::: callout-tip
## Before You Submit

-   [ ] Your name is in the author field at the top of this document
-   [ ] All code chunks run without errors
-   [ ] You've answered all written response questions and shown your work
-   [ ] You've clicked "Render" and successfully generated an HTML file
-   [ ] You're submitting the .qmd file to bCourse
:::