---
title: "Lab 3 Key: Normal Distributions and Z-Scores in R"
author: "Solution Key"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
---

## Load Libraries

```{r}
#| message: false
#| warning: false

library(tidyverse)
```

------------------------------------------------------------------------

## Question 1: Normal Model and Z-Scores (Vehicle Speeds)

**Assume speeds are approximately normal with mean 71 mph and standard deviation 8 mph.**

**a) Compute the z-score for 65 mph and 80 mph.**

```{r}
(65 - 71) / 8
(80 - 71) / 8
```

*Your answer: `z(65) = -0.75`, `z(80) = 1.125`*

**b) Compute `P(speed <= 65)`.**

```{r}
pnorm(65, mean = 71, sd = 8)
```

*Your answer: `0.2266274`*

**c) Compute `P(speed > 80)`.**

```{r}
1 - pnorm(80, mean = 71, sd = 8)
```

*Your answer: `0.1302945`*

**d) Find the speed cutoff where only 10% of vehicles are faster.**

```{r}
qnorm(0.90, mean = 71, sd = 8)
```

*Your answer: `81.25241` mph*

------------------------------------------------------------------------

## Question 2: Z-Score Practice (Heights)

**Assume adult heights are approximately normal with mean 69 inches and standard deviation 3 inches.**

**a) Compute the z-score for a person who is 75 inches tall.**

```{r}
(75 - 69) / 3
```

*Your answer: `2`*

**b) Use `pnorm()` to estimate the percentile for 75 inches.**

```{r}
pnorm(75, mean = 69, sd = 3)
```

*Your answer: `0.9772499` (about the 97.7th percentile)*

**c) Find the 95th percentile height using `qnorm()`.**

```{r}
qnorm(0.95, mean = 69, sd = 3)
```

*Your answer: `73.93456` inches*

**d) Is 75 inches above the 95th percentile?**

*Your answer: Yes. `75 > 73.93456`, so it is above the 95th percentile.*

------------------------------------------------------------------------

## Question 3: Probabilities Above and Between

**Suppose SAT math scores are approximately normal with mean 520 and standard deviation 80.**

**a) Use `pnorm()` to find the proportion of students who score above 640.**

```{r}
1 - pnorm(640, mean = 520, sd = 80)
```

*Your answer: `0.0668072` (about 6.7% of students score above 640)*

**b) Use `pnorm()` to find the proportion who score between 440 and 640.**

```{r}
pnorm(640, mean = 520, sd = 80) - pnorm(440, mean = 520, sd = 80)
```

*Your answer: `0.7745375` (about 77.5% of students score between 440 and 640)*
