---
title: "Lab 3: Normal Distributions and Z-Scores in R"
author: "Your Name Here"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
---

## Setup Instructions

1. Download `lab3.qmd` from bCourse under "labs" > "Lab #3"
2. Place `lab3.qmd` in your `labs` folder. Your folder structure should now look like this:

```
soc106/
├── _quarto.yml
├── data/
│   └── attain.csv
├── assignments/
│   └── hw1.qmd
│   └── hw2.qmd
│   └── hw3.qmd
│   └── hw4.qmd
│   └── hw5.qmd
└── labs/
    ├── lab1.qmd
    ├── lab2.qmd
    └── lab3.qmd
```

3. Use the `Explorer` button on the left to find and open `lab3.qmd`
4. Let's work through it together!

------------------------------------------------------------------------

## Load Libraries

```{r}
#| message: false
#| warning: false

# load libraries
library(tidyverse)
```

------------------------------------------------------------------------

## Function Quick Guide

| Type of question | R function |
|---|---|
| Cumulative probability P(X ≤ x) | `pnorm(x, mean = mu, sd = sigma)` |
| Probability above P(X > x) | `1 - pnorm(x, mean = mu, sd = sigma)` |
| Probability between P(a < X < b) | `pnorm(b, mean, sd) - pnorm(a, mean, sd)` |
| Percentile cutoff | `qnorm(q, mean = mu, sd = sigma)` |
| Standardize a value | `(x - mu) / sigma` |

------------------------------------------------------------------------

## Question 1: Normal Model and Z-Scores (Vehicle Speeds)

Assume speeds are approximately normal with mean 71 mph and standard deviation 8 mph.

1. Compute the z-score for 65 mph and 80 mph.
2. Compute `P(speed <= 65)`.
3. Compute `P(speed > 80)`.
4. Find the speed cutoff where only 10% of vehicles are faster.

```{r}
# Your code here

```

------------------------------------------------------------------------

## Question 2: Z-Score Practice (Heights)

Assume adult heights are approximately normal with mean 69 inches and standard deviation 3 inches.

1. Compute the z-score for a person who is 75 inches tall.
2. Use `pnorm()` to estimate the percentile for 75 inches.
3. Find the 95th percentile height using `qnorm()`.
4. In one sentence: based on your results, is 75 inches above the 95th percentile?

```{r}
# Your code here

```

------------------------------------------------------------------------

## Question 3: Probabilities Above and Between

Suppose SAT math scores are approximately normal with mean 520 and standard deviation 80.

1. Use `pnorm()` to find the proportion of students who score **above 640**.
2. Use `pnorm()` to find the proportion who score **between 440 and 640**.

```{r}
# Your code here

```

------------------------------------------------------------------------

## Wrap-Up

In this lab, you practiced:

- Computing and interpreting z-scores
- Using `pnorm()` to find cumulative normal probabilities
- Using `1 - pnorm()` to find probabilities above a threshold
- Using differences of `pnorm()` to find probabilities between two values
- Using `qnorm()` to find percentile cutoffs

------------------------------------------------------------------------

::: callout-tip
## Before You Leave

- [ ] You ran all code chunks without errors
- [ ] You calculated z-scores by hand in code
- [ ] You used `pnorm()` for cumulative, above, and between probabilities
- [ ] You used `qnorm()` to find a percentile cutoff
- [ ] Your document renders to HTML
:::
