---
title: "Lab 3: Probability Models and Z-Scores in R"
author: "Your Name Here"
date: today
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
---

## Setup Instructions

1. Download `lab3.qmd` from bCourse under "labs" > "Lab #3"
2. Place `lab3.qmd` in your `labs` folder. Your folder structure should now look like this:

```
soc106/
├── _quarto.yml
├── data/
│   └── attain.csv
├── assignments/
│   └── hw1.qmd
│   └── hw2.qmd
│   └── hw3.qmd
│   └── hw4.qmd
│   └── hw5.qmd
└── labs/
    ├── lab1.qmd
    ├── lab2.qmd
    └── lab3.qmd
```

3. Use the `Explorer` button on the left to find and open `lab3.qmd`
4. Let's work through it together!

------------------------------------------------------------------------

## Load Libraries

```{r}
#| message: false
#| warning: false

# load libraries
library(tidyverse)
```

------------------------------------------------------------------------

## Function Quick Guide

| Type of question | Distribution | R function |
|---|---|---|
| Probability of one 0/1 trial | Bernoulli | `dbinom(k, size = 1, prob = pi)` |
| Exact count in n trials | Binomial | `dbinom(k, size = n, prob = pi)` |
| At most / cumulative count | Binomial | `pbinom(k, size = n, prob = pi)` |
| Normal cumulative probability | Normal | `pnorm(x, mean = mu, sd = sigma)` |
| Normal percentile cutoff | Normal | `qnorm(q, mean = mu, sd = sigma)` |
| Standardize a value | Normal | `(x - mu) / sigma` |

------------------------------------------------------------------------

## Question 1: Bernoulli (One Trial)

Suppose a randomly sampled person has a 0.6 probability of voting.

1. Define `pi <- 0.6`.
2. Compute the probability the person voted and the probability they did not vote.
3. Use `rbinom()` to simulate 500 single-person outcomes (`size = 1`) and compute the sample mean.
4. Compare your sample mean to the theoretical mean of a Bernoulli variable.

```{r}
# Your code here

```

------------------------------------------------------------------------

## Question 2: Binomial Exact and Cumulative Probabilities

A player makes free throws with probability 0.75 and takes 8 shots.

1. Compute the probability of making exactly 5 shots.
2. Compute the probability of making at least 7 shots.
3. Compute the probability of making between 4 and 6 shots (inclusive).

Use both `dbinom()` and `pbinom()` where possible.

```{r}
# Your code here

```

------------------------------------------------------------------------

## Question 3: Binomial Mean, SD, and Passing Probability

A student takes a 50-question exam and has an independent 0.8 chance of getting each question correct.

1. Calculate the expected number correct and the standard deviation using formulas.
2. Use R to compute the probability the student passes if passing is at least 35 correct.
3. Simulate 10,000 exams with `rbinom()` and compare the simulated mean to your theoretical mean.

```{r}
# Your code here

```

------------------------------------------------------------------------

## Question 4: Normal Model and Z-Scores (Vehicle Speeds)

Assume speeds are approximately normal with mean 71 mph and standard deviation 8 mph.

1. Compute the z-score for 65 mph and 80 mph.
2. Compute `P(speed <= 65)`.
3. Compute `P(speed > 80)`.
4. Find the speed cutoff where only 10% of vehicles are faster.

```{r}
# Your code here

```

------------------------------------------------------------------------

## Question 5: More Z-Score Practice (Heights)

Assume adult heights are approximately normal with mean 69 inches and standard deviation 3 inches.

1. Compute the z-score for a person who is 75 inches tall.
2. Use `pnorm()` to estimate the percentile for 75 inches.
3. Find the 95th percentile height using `qnorm()`.
4. In one sentence: based on your results, is 75 inches above the 95th percentile?

```{r}
# Your code here

```

------------------------------------------------------------------------

## Wrap-Up

In this lab, you practiced:

- Choosing the right model for the question (Bernoulli, Binomial, Normal)
- Calculating exact and cumulative probabilities in R
- Connecting formulas to simulation
- Calculating and interpreting z-scores
- Finding normal probabilities and percentile cutoffs

------------------------------------------------------------------------

::: callout-tip
## Before You Leave

- [ ] You ran all code chunks without errors
- [ ] You used both `dbinom()` and `pbinom()` at least once
- [ ] You calculated multiple z-scores by hand in code
- [ ] You used both `pnorm()` and `qnorm()`
- [ ] Your document renders to HTML
:::
