---
title: "Lab 5: Hypothesis Testing"
author: "Your Name Here"
date: today
project:
  type: website
  execute-dir: project
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
---

## Overview

In this lab you will run and interpret the four hypothesis tests covered in lecture — the same tests you'll use for hw8. By the end, you should be able to pick the right test for your research question and write up the result.

------------------------------------------------------------------------

## Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(here)

options(scipen = 999)

attain <- read_csv(here("data", "attain.csv"))

attain <- attain |>
  mutate(
    union_member = if_else(union %in% c("r belong", "r and sp"), 1L, 0L),
    college      = if_else(degree %in% c("bachelor", "graduate"), 1L, 0L),
    college_lab  = if_else(college == 1, "College+", "No college")
  )
```

------------------------------------------------------------------------

## Test 1: Proportion z-Test

**Research question:** Is the proportion of U.S. adults with a college degree equal to 25%?

- $H_0: \pi = 0.25$
- $H_1: \pi \neq 0.25$ (two-tailed), $\alpha = 0.05$

```{r}
x <- sum(attain$college, na.rm = TRUE)   # number with college degree
n <- sum(!is.na(attain$degree))          # sample size

prop.test(x, n, p = 0.25, correct = FALSE)
```

::: callout-note
`p = 0.25` sets the **null hypothesis value**. When testing, R uses $SE_0 = \sqrt{\pi_0(1-\pi_0)/n}$ — the null value, not the sample proportion.
:::

**Question 1.** Based on the p-value, do you reject or fail to reject $H_0$? Write one sentence interpreting the result (include the sample proportion and p-value).

*Your answer:*

------------------------------------------------------------------------

## Test 2: One-Sample t-Test

**Research question:** Is the mean years of education among U.S. adults equal to 12 (a high school diploma)?

- $H_0: \mu = 12$
- $H_1: \mu \neq 12$ (two-tailed), $\alpha = 0.05$

```{r}
educ_clean <- attain |> filter(!is.na(educ))

t.test(educ_clean$educ, mu = 12)
```

::: callout-note
`mu = 12` sets the null hypothesis value. The output reports the sample mean, t-statistic, degrees of freedom, p-value, and 95% CI.
:::

**Question 2.** Do you reject or fail to reject $H_0$? Write one sentence interpreting the result (include the sample mean, t-statistic, and p-value).

*Your answer:*

------------------------------------------------------------------------

## Test 3: Two-Sample t-Test

**Research question:** Do married people work a different number of hours per week than never-married people?

- $H_0: \mu_{\text{married}} = \mu_{\text{never married}}$
- $H_1: \mu_{\text{married}} \neq \mu_{\text{never married}}$ (two-tailed), $\alpha = 0.05$

```{r}
married_hrs <- attain |> filter(marital == "married",  !is.na(hrs1)) |> pull(hrs1)
nevmar_hrs  <- attain |> filter(marital == "never ma", !is.na(hrs1)) |> pull(hrs1)

t.test(married_hrs, nevmar_hrs)
```

::: callout-note
The output shows the mean for each group, the t-statistic, and the 95% CI for the **difference** in means. If the CI contains 0, that is consistent with failing to reject $H_0$.
:::

**Question 3.** Do you reject or fail to reject $H_0$? Write 2 sentences reporting the result: include both group means, the t-statistic, the p-value, and your decision.

*Your answer:*

------------------------------------------------------------------------

## Test 4: Chi-Squared Test

**Research question:** Is race associated with having a college degree?

- $H_0$: Race and college attainment are independent
- $H_1$: They are not independent, $\alpha = 0.05$

```{r}
attain_deg <- attain |> filter(!is.na(race), !is.na(degree))

# Contingency table (table() is needed for chisq.test)
tab <- table(attain_deg$race, attain_deg$college_lab)
addmargins(tab)

# College rate within each racial group
attain_deg |>
  group_by(race) |>
  summarize(college_rate = mean(college, na.rm = TRUE))

# Chi-squared test
chisq.test(tab, correct = FALSE)
```

::: callout-note
The chi-squared test tells you **whether** an association exists — the group rates above tell you *where* the differences are.
:::

**Question 4.** Do you reject or fail to reject $H_0$? Write 2 sentences: state your decision (with the $\chi^2$ statistic and p-value), then describe which racial group has the highest/lowest college rate.

*Your answer:*

::: callout-tip
## Before You Leave

-   [ ] You can run all four tests and read the output
-   [ ] You know which test to use based on your variable types
-   [ ] You understand that "fail to reject $H_0$" ≠ "accept $H_0$"
:::
