---
title: "Lab 6: OLS Regression — Answer Key"
author: "Instructor Key"
date: today
project:
  type: website
  execute-dir: project
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
---

## Overview

In this lab you will run a multiple OLS regression using the `attain` dataset. You will use `lm()` to fit the model, `modelsummary()` to display results with robust standard errors, `modelplot()` to visualize coefficients, and residual plots to check OLS assumptions. By the end, you should be able to replicate this workflow on your own research data for hw9.

------------------------------------------------------------------------

## Setup

Load packages, read the data, drop rows with missing values on the variables we will use, and create factor variables with explicit baseline categories.

```{r}
#| message: false
#| warning: false

# load libraries
library(tidyverse)
library(here)
library(modelsummary)

# set scientific notation off
options(scipen = 999)

# load data
attain_raw <- read_csv(here("data", "attain.csv"))

# Drop rows missing any variable that enters the model
attain <- attain_raw |>
  # remove missing data on key variables
  filter(
    !is.na(income91),   # outcome
    !is.na(prestg80),   # continuous predictor
    !is.na(degree),     # categorical predictor
    !is.na(sex)         # binary control
  ) |>
  mutate(
    # Set explicit factor levels so baseline is the first level
    # Baseline for degree = "lt high" (less than high school)
    degree_f = fct_relevel(degree, "lt high", "high sch", "junior c", "bachelor", "graduate"),
    # Baseline for sex = "male"
    sex_f    = fct_relevel(sex, "male")
  )

# How many rows did we drop?
cat("Original rows:", nrow(attain_raw), "\n")
cat("Rows after dropping missing:", nrow(attain), "\n")
cat("Rows dropped:", nrow(attain_raw) - nrow(attain), "\n")
```

::: callout-note
We drop missing values in the data-loading step so that all models are estimated on the **same sample**. If you filter inside `lm()` or across different steps, sample sizes can vary across models, making comparisons misleading.
:::

------------------------------------------------------------------------

## Step 1 — Research Question and Variables

We will estimate the following regression:

$$\widehat{\text{income}} = \alpha + b_1\text{prestige} + b_2\text{degree} + b_3\text{sex} + e$$

| Role | Variable | Description |
|------------------------|------------------------|------------------------|
| **Outcome (Y)** | `income91` | Household income in dollars — continuous |
| **Key predictor** | `prestg80` | Occupational prestige score — continuous |
| **Categorical control** | `degree_f` | Educational degree — 5 levels, baseline = *lt high* |
| **Binary control** | `sex_f` | Sex — binary, baseline = *male* |

**Our research question:** After controlling for educational attainment and sex, how does occupational prestige impact household income?

------------------------------------------------------------------------

## Step 2 — Visualize Your Variables First

Always look at your data before running a regression.

**Continuous predictor vs. outcome:**

```{r}
#| message: false
#| warning: false

ggplot(attain, aes(x = prestg80, y = income91)) +
  geom_point(alpha = 0.2, color = "gray50", size = 1) +
  geom_smooth(method = "lm", se = TRUE,
              color = "#4E79A7", fill = "#4E79A7", alpha = 0.2) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(title = "Occupational Prestige and Household Income",
       x = "Occupational Prestige Score",
       y = "Household Income ($)") +
  theme_minimal()
```

**Categorical predictor vs. outcome:**

```{r}
attain |>
  group_by(degree_f) |>
  summarize(mean_income = mean(income91, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = degree_f, y = mean_income)) +
  geom_col(fill = "#4E79A7") +
  scale_y_continuous(labels = scales::dollar_format(),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Mean Household Income by Educational Degree",
       x = "Degree (left = lowest; right = highest)",
       y = "Mean Income") +
  theme_minimal()
```

**Question 1.** Looking at the two plots above, describe the direction of each relationship in one sentence each. Does income go up or down as prestige increases? How does income differ across degree levels?

*Answer:* Income increases as occupational prestige increases — the scatter plot shows a positive upward-sloping trend, confirmed by the fitted regression line. Income rises stepwise across degree levels, with the largest jumps at the higher end of the education distribution (bachelor's and graduate degrees).

------------------------------------------------------------------------

## Step 3 — Fit the OLS Model

### Why This Model?

Every regression requires a theoretical justification for which variables belong in the equation. Here is the logic behind each choice:

-   **Outcome — `income91`**: We are interested in economic inequality. Household income is a direct and continuous measure of material well-being, making OLS appropriate.
-   **Key predictor — `prestg80`**: Occupational prestige captures social standing and job desirability. The core question is whether higher-prestige occupations translate to higher earnings *after* accounting for the credentials workers bring to those jobs and gender differences in pay.
-   **Control — `degree_f`**: Education is a strong driver of both occupational placement and earnings. If we omit it, any apparent prestige effect may really reflect the fact that more educated people hold more prestigious *and* higher-paying jobs — a classic confounder.
-   **Control — `sex_f`**: A well-documented gender pay gap exists in the U.S. labor market, and women are also sorted into lower-prestige occupations on average. Controlling for sex isolates the prestige–income relationship from those gender differences.

Including both controls means the prestige coefficient tells us: *among people with the same education level and sex, how does a one-unit difference in occupational prestige impact income?*

```{r}
model <- lm(income91 ~ prestg80 + degree_f + sex_f, data = attain)

summary(model)
```

::: callout-note
`summary()` gives you the raw output. You can read the coefficient estimates, standard errors, t-values, and p-values here — but we will use `modelsummary()` below for a cleaner table. Note that the SEs from `summary()` are **classical** (not robust) — we'll switch to robust SEs in Step 4.
:::

------------------------------------------------------------------------

## Step 4 — Regression Table with Robust Standard Errors

Use `modelsummary()` with `vcov = "HC1"` to display results with **heteroskedasticity-robust standard errors**. This is best practice when working with income data, which often shows unequal variance across the range of fitted values.

```{r}
modelsummary(
  model,
  vcov      = "HC1",
  coef_rename = c(
    "(Intercept)"        = "Intercept",
    "prestg80"           = "Occupational Prestige",
    "degree_fhigh sch"   = "High School (vs. lt high)",
    "degree_fjunior c"   = "Junior College (vs. lt high)",
    "degree_fbachelor"   = "Bachelor's Degree (vs. lt high)",
    "degree_fgraduate"   = "Graduate Degree (vs. lt high)",
    "sex_ffemale"        = "Female (vs. male)"
  ),
  stars     = TRUE,
  gof_map   = c("nobs", "r.squared", "adj.r.squared"),
  title     = "OLS: Household Income ~ Prestige + Degree + Sex (HC1 Robust SEs)"
)
```

::: callout-note
**What `vcov = "HC1"` does:** It replaces the classical standard errors with robust ones that are valid even if the residual variance is not constant across fitted values (heteroskedasticity). The **coefficient estimates do not change** — only the standard errors, t-statistics, and p-values.
:::

------------------------------------------------------------------------

## Step 5 — Visualize Coefficients

A coefficient plot shows each estimate with its 95% confidence interval. If the CI **does not cross zero**, the coefficient is statistically significant at α = 0.05.

```{r}
#| fig-height: 3.5

modelplot(
  model,
  vcov      = "HC1",
  coef_omit = "Intercept",
  coef_rename = c(
    "prestg80"           = "Occupational Prestige",
    "degree_fhigh sch"   = "High School (vs. lt high)",
    "degree_fjunior c"   = "Junior College (vs. lt high)",
    "degree_fbachelor"   = "Bachelor's Degree (vs. lt high)",
    "degree_fgraduate"   = "Graduate Degree (vs. lt high)",
    "sex_ffemale"        = "Female (vs. male)"
  )
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#E15759") +
  labs(
    title = "OLS Coefficients with 95% Confidence Intervals (HC1 Robust SEs)",
    x     = "Estimated Coefficient ($)"
  ) +
  theme_minimal()
```

**Question 2.** Look at the coefficient plot. Which coefficients are statistically significant (CI does not cross the dashed red line at zero)? List them.

*Answer:* The following coefficients are statistically significant: Occupational Prestige, Bachelor's Degree (vs. lt high), Graduate Degree (vs. lt high), and Female (vs. male). High School and Junior College dummies may or may not cross zero — check your specific output. The prestige CI may look visually narrow relative to the degree gaps; see the note below for why this is a scale artifact rather than evidence of non-significance.

::: callout-note
**Why does the prestige CI look so narrow compared to the degree gaps?**

This is a scale artifact, not a substantive finding. The prestige score runs roughly 17–86, so a *1-point* change is a tiny increment. The degree dummies, by contrast, compare entire education tiers against the baseline, producing large dollar gaps that dominate the x-axis. Both are estimated on the same dollar scale, so the prestige CI appears narrow even if it is far from zero.

**Two ways to deal with this:**

1.  **Check the table directly.** The stars and p-values in Step 4 are the authoritative test. A visually narrow CI on a shared-scale plot does not mean a coefficient is insignificant.
2.  **Rescale the predictor.** Replace `prestg80` with `I(prestg80 / 10)` inside `lm()`. Now the coefficient represents income change per *10-point* increase in prestige — a more meaningful unit — and the CI will visually stretch on the plot.

**Back-of-napkin significance check — the "2× SE rule":** If \|coefficient\| ≥ 2 × SE, it is approximately significant at α = 0.05. This works because the critical t-value at α = 0.05 is ≈ 1.96 ≈ 2, so t = b / SE ≥ 2 implies p ≤ 0.05. Check this against the prestige row in your Step 4 table: multiply the SE by 2 and see if the coefficient is larger. It should be — by a lot.
:::

------------------------------------------------------------------------

## Step 6 — Interpret Your Results

Use the regression table from Step 4 to answer the following questions. Write full sentences.

**Question 3 — Continuous predictor.** Interpret the coefficient on `Occupational Prestige`. Use this template:

> "For every one unit increase in occupational prestige, predicted household income \[increases/decreases\] by \$\_\_\_, holding educational degree and sex constant."

*Answer:* For every one unit increase in occupational prestige, predicted household income increases by \$\[value from your output\], holding educational degree and sex constant.

::: callout-note
The prestige score ranges from roughly 17 to 86 — a span of about 70 points. To get a sense of the practical magnitude, multiply the per-point coefficient by a realistic range (e.g., 10 or 20 points) rather than interpreting a single-point change in isolation.
:::

------------------------------------------------------------------------

**Question 4 — Categorical predictor.** Interpret the coefficient on `Bachelor's Degree (vs. lt high)`. Remember: the baseline category is *less than high school*. Use this template:

> "Compared to respondents with less than a high school degree, those with a bachelor's degree are predicted to earn \$\_\_\_ \[more/less\] per year, holding prestige and sex constant."

*Answer:* Compared to respondents with less than a high school degree, those with a bachelor's degree are predicted to earn \$\[value from your output\] more per year, holding prestige and sex constant.

------------------------------------------------------------------------

**Question 5 — Binary predictor.** Interpret the coefficient on `Female (vs. male)`. Remember: the baseline is *male*. Use this template:

> "Compared to men, women are predicted to earn \$\_\_\_ \[more/less\] per year, holding prestige and degree constant."

*Answer:* Compared to men, women are predicted to earn \$\[value from your output\] less per year, holding prestige and degree constant. (The coefficient on Female is negative, reflecting a persistent gender pay gap even after controlling for occupational prestige and educational attainment.)

------------------------------------------------------------------------

**Question 6 — R².** Report the R² value from the table and interpret it in one sentence. What proportion of variation in income does this model explain?

*Answer:* The R² value is \[value from your output\]. This model explains approximately \[R² × 100\]% of the variation in household income. The remaining variation is attributable to factors not included in the model, such as work experience, hours worked, industry, or geographic region.

------------------------------------------------------------------------

**Question 7 — Statistical significance.** Look at the stars on the `Occupational Prestige` coefficient. Is this coefficient statistically significant at the α = 0.05 level? What does that mean — do you reject or fail to reject $H_0: \beta_{\text{prestige}} = 0$?

*Answer:* Yes, the Occupational Prestige coefficient is statistically significant at the α = 0.05 level (indicated by the star(s) in the table and a p-value below 0.05). We reject the null hypothesis $H_0: \beta_{\text{prestige}} = 0$. This means the observed relationship between prestige and income is unlikely to be due to chance alone, given the controls for degree and sex. We conclude there is a statistically significant positive association between occupational prestige and household income.

------------------------------------------------------------------------

## Step 7 — Check OLS Assumptions

### Residuals vs. Fitted (linearity + homoskedasticity)

If the relationship is linear and variance is constant, the points should be randomly scattered around the horizontal line with no clear pattern.

```{r}
#| message: false
#| warning: false

attain |>
  mutate(
    fitted = fitted(model),
    resid  = residuals(model)
  ) |>
  ggplot(aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.2, color = "gray40", size = 1) +
  geom_hline(yintercept = 0, color = "#E15759", linetype = "dashed", linewidth = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "#4E79A7", linewidth = 1) +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Residuals vs. Fitted Values",
    x     = "Fitted Values (Predicted Income)",
    y     = "Residuals"
  ) +
  theme_minimal()
```

**Question 8.** Does the blue loess curve look approximately flat (suggesting linearity is satisfied)? Does the vertical spread of points look roughly constant across fitted values, or does it widen? What does this suggest about heteroskedasticity?

*Answer:* The loess curve is approximately flat, suggesting the linearity assumption is reasonably satisfied — there is no strong systematic curvature in the residuals. However, the vertical spread of points widens noticeably at higher fitted values (a fan or funnel shape), which is evidence of **heteroskedasticity**: the residual variance is not constant across the range of predictions. This is the expected pattern for income data, where high-income households exhibit more variability. This is exactly why we used `vcov = "HC1"` robust standard errors in Step 4 — they remain valid under heteroskedasticity.

------------------------------------------------------------------------

### Histogram of Residuals (normality of errors)

If the normality assumption holds, residuals should look roughly bell-shaped. In large samples this matters less (CLT), but it is good practice to check.

```{r}
tibble(resid = residuals(model)) |>
  ggplot(aes(x = resid)) +
  geom_histogram(bins = 40, fill = "#4E79A7", color = "white") +
  labs(
    title = "Distribution of Residuals",
    x     = "Residual ($)",
    y     = "Count"
  ) +
  theme_minimal()
```

**Question 9.** Does the histogram of residuals look approximately normal (symmetric, bell-shaped), or is it skewed? If skewed, what transformation of the outcome variable might help?

*Answer:* The histogram of residuals is **right-skewed** (a long right tail), not symmetric. This reflects the underlying skewness of income distributions — most respondents cluster at lower to middle incomes, but a small number earn very high incomes, pulling the distribution and its residuals to the right. A **log transformation** of the outcome variable — replacing `income91` with `log(income91)` in the `lm()` call — would likely produce more normally distributed residuals, since log-income is approximately normal in most survey datasets. Note: in large samples, the Central Limit Theorem means non-normal errors have minimal impact on the validity of inference, but the transformation would also address the heteroskedasticity seen in Question 8.

------------------------------------------------------------------------

::: callout-tip
## Putting It Together

Even if the residuals vs. fitted plot shows some heteroskedasticity, our results are still valid because we used `vcov = "HC1"` robust standard errors in Step 4. The coefficients themselves are unaffected by heteroskedasticity — only the standard errors are, and HC1 corrects for that.
:::

------------------------------------------------------------------------

::: callout-tip
## Before You Leave

-   [ ] I can load data, drop missing values, and create factor variables with a specified baseline
-   [ ] I can fit an OLS model with `lm()` and read the `summary()` output
-   [ ] I can produce a clean regression table with robust SEs using `modelsummary(vcov = "HC1")`
-   [ ] I can visualize coefficients with `modelplot()` and interpret whether CIs cross zero
-   [ ] I can interpret a continuous coefficient, a categorical coefficient, and a binary coefficient
-   [ ] I can produce and interpret a residuals vs. fitted plot and a residual histogram
:::