---
title: "Weekly Assignment #1"
author: "Your Name Here"  # Replace with your name
date: "02/05/2026"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
---


::: callout-important
## Submission Instructions

1.  Download this `hw1.qmd` file. The `attain.csv` data should already be in your data folder.
2.  Save the `hw1.qmd` file in the assignments folder.
3.  Complete all questions by writing your code in the provided code chunks. Replace `# Your code here` lines with your actual code.
4.  Click the "Render" button at the top of Positron to generate an HTML file.
5.  Submit **BOTH** the .qmd file AND the .html file to bCourse before the deadline on February 5, 2026
6.  Make sure your name appears in the "author" field in the YAML header above.
7. Make sure your document renders without errors before submitting.
:::

Once you add `hw1.qmd`, your folder structure should look like this:

```
soc106/
├── _quarto.yml
├── assignments/
│   └── hw0.qmd
│   └── hw1.qmd
├── data/
│   └── attain.csv
└── labs/
    └── lab1.qmd
```


## Setup 

Run this code to set up your environment. If you need to install any libraries, make sure to do that in the console.

```{r}
# Clear environment
# ---------
rm(list = ls())

# install libraries -- These should already have been installed but they are here just in case.

# ---------
# install.packages("tidyverse")
# install.packages("here")

# load libraries
# ---------
library(tidyverse)
library(here)
```

## Question 1

Load the data file `attain.csv` into R using the `read_csv` function. Be sure to use the `here()` library to ensure you are using a relative file path and that the csv file is saved in the "data" folder and the .qmd file is saved in the "assignments" folder. Once you've loaded the dataset, use the `glimpse()` function to examine the structure of the data and the variables.

```{r}
#| warning: false

# Your code here

```

**Write 1-2 sentences describing what you observe in the data:**

*Your response here*

------------------------------------------------------------------------

## Question 2

Create a new object called `ages` from the `attain` dataset by selecting the column name called `age` using the  appropriate tidyverse function and glimpsing the object. What is the type of variable in it? (e.g., numeric, character, boolean). Specifically, what is this called?

```{r}

# Your code here

```

**What is the type of variable?**

*Your response here*

------------------------------------------------------------------------

## Question 3

Now create a new object from the the original `attain` dataset (column name = `marital`), call it `marital_status`, and view the object using the `print` function. What is the type of variable in this column vector?

```{r}

# Your code here

```

**What is the type of variable?**

*Your response here*

------------------------------------------------------------------------

## Question 4

Suppose we want to only look at a smaller dataset that contains the respondent's age, sex, race, marital status, and age at marriage (column names `age`, `sex`, `race`, `marital`, `agewed`, respectively). Create a subset of the original `attain` dataset (call it `subset1`) using the appropriate tidyverse function. Make sure to glimpse it so we can see what it looks like.

```{r}

# Your code here

```

------------------------------------------------------------------------

## Question 5

Now suppose we want a smaller dataset that only contains the first 1000 observations. Create a subset of `attain` data (call it `subset2`) using the appropriate tidyverse function. Make sure to glimpse it. How do you know you did it correctly?

```{r}

# Your code here

```

**How do you know you did it correctly?**

*Your response here*

------------------------------------------------------------------------

## Question 6

Now, suppose we are really concerned about unemployment within our sample (column name `unemp`). But, we notice that many respondents have missing unemployment statuses (i.e., `unemp` is equal to `NA`). Let's create a subset that consists of all respondents who do not have missing unemployment information. (*Hint: to find observations that are not missing, use the `!` (not) logical operator and the `is.na` function.*)

```{r}

# Your code here
# Don't forget to check how many respondents are in your subset!

```

**How many respondents are in this subset?**

*Your response here*

------------------------------------------------------------------------

## Question 7

Repeat Question 6 (call it `subset4`), but this time creating a subset of respondents who all have information about their father's education (column name `paeduc`).

```{r}

# Your code here

```

**How many respondents are in this subset?**

*Your response here*

------------------------------------------------------------------------

## Question 8

Now suppose we want to investigate the relationship between unemployment status and father's education. This time, create a subset (call it `subset5`) of respondents who have non-missing information for **both** unemployment status and father's education. (*Hint: You will need to use a logical operator again for this one.*)

```{r}

# Your code here

```

**How many respondents are in this subset?**

*Your response here*

------------------------------------------------------------------------

## Question 9

Create a new variable in the dataset that measures the number of years since the birth of the respondent's first child, using the variables `age` and `agekdbrn`. Call the new dataset `attain_kids`. Use the appropriate tidyverse function to do this (*Hint: Create a new variable and call it `years_since_kid`. You'll have to do some subtration here!*)

```{r}

# Your code here

```

------------------------------------------------------------------------

## Question 10

Finally, there is a left over variable in this dataset from the original .csv file that we want to drop. Using the `attain` dataset again, let's' first rename the variable--call it "to_drop"--and then drop it from the dataset (*Hint: You can use a negative sign before the variable to select everything but that one.*). How can you tell for sure that this worked (*Hint: think about how many columns you should have.*)

```{r}

# Your code here

```

**How can you tell this worked?**

*Your response here*

------------------------------------------------------------------------

::: callout-tip
## Before You Submit

-   [ ] Your name is in the author field at the top of this document
-   [ ] All code chunks run without errors
-   [ ] You've answered all written response questions
-   [ ] You've clicked "Render" and successfully generated an HTML file
-   [ ] You're submitting BOTH the .qmd AND .html files to bCourse
:::
